---
title: "figures"
output: html_document
date: "2023-08-15"
---

```{r setup, include=FALSE}
library(ggplot2)

```


```{r scripts and dirs}

workDir <- gitDir <- "~/Documents/giterdone/"

dataDir <- paste0(workDir,"data/")
figDir <- paste0(workDir,"figures/")
outDir <- paste0(workDir,"output/")

dir.create(dataDir)
dir.create(figDir)
dir.create(outDir)
```


```{r setup, include=FALSE}
library(ggplot2)
library(qvalue)
library(stringr)
library(edgeR)
library(EnvStats)
library(ggbeeswarm)
library(UpSetR)

workDir <- "~/Dropbox (Personal)/meerkats/RNA/meerkatGE/"
dataDir <- paste0(workDir,"data/")
figDir <- paste0(workDir,"figures/")
outDir <- paste0(workDir,"output/")
mashDir <- paste0(workDir,"mash/")

figPaperDir <- paste0(figDir,"paper/")
dir.create(figPaperDir)
suppFigDir <- paste0(figDir,"paper/supp/")
dir.create(suppFigDir)

# COLORS #

#dark=m, light=f
#Ctrl
ctrlM <- "grey40"
ctrlF <- "grey70"
ctrlF <- "grey20"
ctrlF <- ""
ctrlM <- "darkorchid4"
ctrlM <- "#B5A9F4"
ctrlF <- "#154376"
ctrlM <- "#4B9CD3"

#Gard
gardF <- "#947322"
gardM <- "#E0B13F"

#LPS
lpsF <- "#944F40"
lpsM <- "#E06E55"

#Dex
dexM <- "#4BC4E0"
dexF <- "#358394"


highES <- "forestgreen"
lowES <- "orange"


#LRS
lowLRS <- "deeppink"
highLRS <- "darkorchid"
highLRS <- "#779CD7"
lowLRS <- "#E08FC6"


#hallmark
gene.set <- read.table(paste0(dataDir,"hallmarkPathways_wCategories.csv"), sep = ",")
colnames(gene.set) <- c("Associated.Gene.Name","GO.Term.Accession","Category")

```

Figure 1
```{r 1.A}
allMetaF <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_allTrt_10000genes_5lCPM_metadata.RDS")
allMetaM <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_allTrt_10000genes_5lCPM_metadata.RDS")

#capture only
allMetaF <- unique(allMetaF[colnames(allMetaF) %in% c("group","captureref","sex","status_dom_sub",
                                                      "ageAtSample","evidenceofTB.","draw_time","bloodVolmL",
                                                      "sampleDate","sampleYear","sampleMonth","rationale","transitionID",
                                                      "correctsampleID","birthdate","collectionRationale","climSeason")])
allMetaF$sMonthDec <- round(as.numeric(format(as.Date(allMetaF$sampleDate, format="%Y-%m-%d"),"%m")) +
                              as.numeric(format(as.Date(allMetaF$sampleDate, format="%Y-%m-%d"),"%d")) / 31, digits = 2)

allMetaM <- unique(allMetaM[colnames(allMetaM) %in% c("group","captureref","sex","status_dom_sub",
                                                      "ageAtSample","evidenceofTB.","draw_time","bloodVolmL",
                                                      "sampleDate","sampleYear","sampleMonth","rationale","transitionID",
                                                      "correctsampleID","birthdate","collectionRationale","climSeason")])
allMetaM$sMonthDec <- round(as.numeric(format(as.Date(allMetaM$sampleDate, format="%Y-%m-%d"),"%m")) +
                              as.numeric(format(as.Date(allMetaM$sampleDate, format="%Y-%m-%d"),"%d")) / 31, digits = 2)


allMeta <- rbind.data.frame(allMetaF,allMetaM)


ggplot(allMetaF, aes(x = sMonthDec)) +
  geom_density(fill = ctrlF) +
  theme_minimal() + theme_bw() +
  theme(axis.text.y = element_text(size = 25))
ggsave(paste0(figPaperDir,"1a_densityF.png"), width = 8, height = 4)

ggplot(allMetaM, aes(x = sMonthDec)) +
  geom_density(fill = ctrlM) +
  theme_minimal() + theme_bw() +
  theme(axis.text.y = element_text(size = 25))
ggsave(paste0(figPaperDir,"1a_densityM.png"), width = 8, height = 4)

ggplot(allMetaF, aes(x = sMonthDec)) +
  geom_histogram(fill = ctrlF, bins = 12) +
  theme_minimal() + theme_bw() +
  theme(axis.text.y = element_text(size = 25)) +
  ylim(c(0,25))
ggsave(paste0(figPaperDir,"1a_histF.png"), width = 8, height = 4)

ggplot(allMetaM, aes(x = sMonthDec)) +
  geom_histogram(fill = ctrlM, bins = 12) +
  theme_minimal() + theme_bw() +
  theme(axis.text.y = element_text(size = 25)) +
  ylim(c(0,25))
ggsave(paste0(figPaperDir,"1a_histM.png"), width = 8, height = 4)

```
```{r 1.B longitudinal}
allMetaF <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_allTrt_10000genes_5lCPM_metadata.RDS")
allMetaM <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_allTrt_10000genes_5lCPM_metadata.RDS")

#capture only
allMetaF <- unique(allMetaF[colnames(allMetaF) %in% c("group","captureref","sex","status_dom_sub",
                                                      "ageAtSample","evidenceofTB.","draw_time","bloodVolmL",
                                                      "sampleDate","sampleYear","sampleMonth","rationale","transitionID",
                                                      "correctsampleID","birthdate","collectionRationale","climSeason")])
allMetaF$sMonthDec <- round(as.numeric(format(as.Date(allMetaF$sampleDate, format="%Y-%m-%d"),"%m")) +
                              as.numeric(format(as.Date(allMetaF$sampleDate, format="%Y-%m-%d"),"%d")) / 31, digits = 2)

allMetaM <- unique(allMetaM[colnames(allMetaM) %in% c("group","captureref","sex","status_dom_sub",
                                                      "ageAtSample","evidenceofTB.","draw_time","bloodVolmL",
                                                      "sampleDate","sampleYear","sampleMonth","rationale","transitionID",
                                                      "correctsampleID","birthdate","collectionRationale","climSeason")])
allMetaM$sMonthDec <- round(as.numeric(format(as.Date(allMetaM$sampleDate, format="%Y-%m-%d"),"%m")) +
                              as.numeric(format(as.Date(allMetaM$sampleDate, format="%Y-%m-%d"),"%d")) / 31, digits = 2)


allMeta <- rbind.data.frame(allMetaF,allMetaM)

caughtMulti <- names(table(allMeta$correctsampleID)[which(table(allMeta$correctsampleID) > 1)])

caughtMultiStatus <- vector()

for (s in caughtMulti) {
  statuses <- unique(subset(allMeta, correctsampleID == s)$status_dom_sub)
  if (length(statuses) > 1) {
    caughtMultiStatus <- c(caughtMultiStatus,s)
  }
}

caughtMultiStatus <- caughtMultiStatus[which(caughtMultiStatus != "VVHF129")]

multiMeta <- subset(allMeta, correctsampleID %in% caughtMultiStatus)

### sort by sex
factorOrder <- unique(multiMeta$correctsampleID[order(multiMeta$sex)])
multiMeta$correctsampleID <- factor(multiMeta$correctsampleID, levels = factorOrder)


ggplot(data = multiMeta,
       aes(x = sampleDate,
           y = correctsampleID,
           group = correctsampleID,
           col = sex,
           shape = status_dom_sub)) +
  geom_point() +
  geom_line()


### sort by earliest capture
factorOrder <- unique(multiMeta$correctsampleID[order(multiMeta$sampleDate)])
multiMeta$correctsampleID <- factor(multiMeta$correctsampleID, levels = factorOrder)


ggplot(data = multiMeta,
       aes(x = sampleDate,
           y = correctsampleID,
           group = correctsampleID,
           col = sex,
           shape = status_dom_sub)) +
  geom_point(size = 4) +
  geom_line(lwd = 1) +
  theme_minimal() + theme_bw() +
  theme(axis.text.y = element_text(size = 14),
        axis.title = element_text(color = "black"),
        axis.text = element_text(color = "black"),
        legend.position = "none") +
  scale_shape_manual(values = c(16, 1)) +
  scale_color_manual(values = c(ctrlF, ctrlM))
ggsave(paste0(figPaperDir,"1B_LongitudinalSampling_black.png"), width = 12, height = 4)
```


```{r 1.B}
#pca of responses, LPS/Gard/Dex

pointSize <- 3
plotAlpha <- 1
labelSize <- 20
numSize <- 16

pcaLPS <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_MF_LPS_10000genes_5lCPM_modelF_dfMeta.RDS")

pcaLPS$TrtSex <- paste0(pcaLPS$Trt,"_",pcaLPS$sex)

ggplot(data = pcaLPS,
       aes(x = PC1,
           y = PC2,
           col = TrtSex)) +
  geom_point(size = pointSize, alpha = plotAlpha) +
  theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = numSize, color = "black"),
        axis.title = element_text(size = labelSize, color = "black"),
        panel.grid = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c(ctrlF,ctrlM,lpsF,lpsM))
ggsave(paste0(figPaperDir,"1b_LPS.png"), width = 4, height = 4)


pcaGard <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_MF_Gard_10000genes_5lCPM_modelG_dfMeta.RDS")

pcaGard$TrtSex <- paste0(pcaGard$Trt,"_",pcaGard$sex)

ggplot(data = pcaGard,
       aes(x = PC1,
           y = PC2,
           col = TrtSex)) +
  geom_point(size = pointSize, alpha = plotAlpha) +
  theme_minimal() + theme_bw() +
    theme(axis.text = element_text(size = numSize, color = "black"),
        axis.title = element_text(size = labelSize, color = "black"),
        panel.grid = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c(ctrlF,ctrlM,gardF,gardM))
ggsave(paste0(figPaperDir,"1b_Gard.png"), width = 4, height = 4)


pcaDex <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_MF_Dex_10000genes_5lCPM_modelG_dfMeta.RDS")

pcaDex$TrtSex <- paste0(pcaDex$Trt,"_",pcaDex$sex)

ggplot(data = pcaDex,
       aes(x = PC1,
           y = PC2,
           col = TrtSex)) +
  geom_point(size = pointSize, alpha = plotAlpha) +
  theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = numSize, color = "black"),
        axis.title = element_text(size = labelSize, color = "black"),
        panel.grid = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c(ctrlF,ctrlM,dexF,dexM))

ggsave(paste0(figPaperDir,"1b_Dex.png"), width = 4, height = 4)
#print R correlations

```

```{r Methods Text}
allMeta <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_MF_allTrt_10000genes_5lCPM_metadata.RDS")

totalWeights <- unique(allMeta[,colnames(allMeta) %in% c("captureref","totalWeights")])

paste0("We measured meerkat body mass multiple times daily, with the samples in this analysis averaging of ",
       round(mean(totalWeights$totalWeights), digits = 1),
       " weighings in the 60 days surrounding the date of sampling")

ages <- unique(allMeta[,colnames(allMeta) %in% c("captureref","ageAtSample")])
paste0("The average age at the date of sampling was ",
       round(mean(ages$ageAtSample), digits = 1),
       " years (sd=",round(sd(ages$ageAtSample), digits = 2),
       " years; ages ",min(ages$ageAtSample),
       "-",max(ages$ageAtSample),
       ")")

massLPS <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_weight")

print(paste0("Body mass exhibited a modest correlation with gene expression in the LPS-stimulated condition (",sum(massLPS$fdr_weight < .1),
             " body mass-associated genes; FDR < 10%)-"))

cor.test(massLPS$beta_weight, massLPS$beta_status)$estimate
cor.test(massLPS$beta_weight, massLPS$beta_status)$p.value
print(paste0("but this signature was substantially weaker than that of dominance status itself (rmass-status = ",
             cor.test(massLPS$beta_weight, massLPS$beta_status)$estimate,", p=",
             cor.test(massLPS$beta_weight, massLPS$beta_status)$p.value,")"))

ggplot(data = massLPS,
       aes(x = beta_status,
           y = beta_weight)) +
  geom_point()

massweightCor <- cor.test(massLPS$beta_weight, massLPS$beta_status)


print(paste0("(rage-status = ",cor.test(massLPS$beta_age, massLPS$beta_status)$estimate,", p=",
             cor.test(massLPS$beta_age, massLPS$beta_status)$p.value,", rpregnancy-status = ",
             cor.test(massLPS$beta_preg, massLPS$beta_status)$estimate,", p=",
             cor.test(massLPS$beta_preg, massLPS$beta_status)$p.value,""))


massAllTrt <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_allTrt_10000genes_5lCPM_modelG_allTrtF+Preg+WeightwKmatrix_emmreml")

print(paste0("(rage-status = ",cor.test(massAllTrt$beta_age, massAllTrt$beta_status)$estimate,", p=",
             cor.test(massAllTrt$beta_age, massAllTrt$beta_status)$p.value,", rpregnancy-status = ",
             cor.test(massAllTrt$beta_preg, massAllTrt$beta_status)$estimate,", p=",
             cor.test(massAllTrt$beta_preg, massAllTrt$beta_status)$p.value,""))
```


```{r 1.C}
### 1C
#count of significant genes, male and female by treatment | using qvalue as a dummy
maleLPS <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_LPS_10000genes_5lCPM_modelF_basic+WeightwKmatrix_emmreml_wPermFDR_v1_treatment")
#maleLPS$qval_treatment <- qvalue(maleLPS$pval_treatment)$qvalue

femaleLPS <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_treatment")
#femaleLPS$qval_treatment <- qvalue(femaleLPS$pval_treatment)$qvalue


maleGard <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_Gard_10000genes_5lCPM_modelG_basic+WeightwKmatrix_emmreml_wPermFDR_v1_treatment")
#maleGard$qval_treatment <- qvalue(maleGard$pval_treatment)$qvalue

femaleGard <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Gard_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_treatment")
#femaleGard$qval_treatment <- qvalue(femaleGard$pval_treatment)$qvalue


maleDex <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_Dex_10000genes_5lCPM_modelG_basic+WeightwKmatrix_emmreml_wPermFDR_v1_treatment")
#maleDex$qval_treatment <- qvalue(maleDex$pval_treatment)$qvalue

femaleDex <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Dex_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_treatment")
#femaleDex$qval_treatment <- qvalue(femaleDex$pval_treatment)$qvalue


fdrThresh <- 0.10

trtSigGeneVector <- c(rep("mLPS",sum(maleLPS$fdr_treatment < fdrThresh)),
                      rep("fLPS",sum(femaleLPS$fdr_treatment < fdrThresh)),
                      rep("mGard",sum(maleGard$fdr_treatment < fdrThresh)),
                      rep("fGard",sum(femaleGard$fdr_treatment < fdrThresh)),
                      rep("mDex",sum(maleDex$fdr_treatment < fdrThresh)),
                      rep("fDex",sum(femaleDex$fdr_treatment < fdrThresh)))

trtSigGeneDF <- as.data.frame(trtSigGeneVector)

trtSigGeneDF$trtSigGeneVector <- factor(trtSigGeneDF$trtSigGeneVector, levels = c("mDex","fDex","mGard","fGard","mLPS","fLPS"))

ggplot() +
  geom_bar(dat = trtSigGeneDF,
       aes(x = trtSigGeneVector,
           fill = trtSigGeneVector), width = .35) +
  geom_point(aes(x = names(table(trtSigGeneVector)),
                 y = table(trtSigGeneVector),
                 col = names(table(trtSigGeneVector))), size = 4.5) +
  theme_minimal() + theme_bw() +
  theme(axis.text.y = element_text(size = 14),
        panel.grid = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
  scale_fill_manual(values = c(dexM,dexF,gardM,gardF,lpsM,lpsF)) +
  scale_color_manual(values = c(dexF,gardF,lpsF,dexM,gardM,lpsM)) +
  theme(legend.position = "none")

ggsave(paste0(figPaperDir,"1c_barplot.png"), width = 2.5, height = 8)
```


```{r 1.D}
### 1D
#enrichment scores of treatment across male and females
gseaF <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/gsea_trtstatuspreg_GSEA_allResults.txt", 
                    sep = ",", header = TRUE)
colnames(gseaF)[1] <- "pathway"
colnames(gseaF)[4] <- "ESup"
colnames(gseaF)[5] <- "ESdown"

#add category
gseaF$cat <- ""

for (row in 1:dim(gseaF)[1]) {
  pathCat <- subset(gene.set, GO.Term.Accession == gseaF$pathway[row])$Category[1]
  gseaF$cat[row] <- pathCat 
}

gseaF$sigES <- ifelse(gseaF$pval_upreg < gseaF$pval_downreg, gseaF$ESup, gseaF$ESdown)
gseaF$sigPval <- ifelse(gseaF$pval_upreg < gseaF$pval_downreg, gseaF$pval_upreg, gseaF$pval_downreg)

gseaF$biggerES <- ifelse(abs(gseaF$ESup) > abs(gseaF$ESdown), gseaF$ESup, gseaF$ESdown)

trtGSEA <- subset(gseaF, c == "treatment")
#trtGSEA <- subset(gseaF, c == "status")

immunePaths <- unique(subset(gseaF, cat == "immune")$pathway)
immunePaths <- immunePaths[-which(immunePaths == "HALLMARK_INFLAMMASOME")]
immunePaths <- c(immunePaths,"HALLMARK_TNFA_SIGNALING_VIA_NFKB")

trtGSEAimmune <- subset(trtGSEA, pathway %in% immunePaths)

trtGSEAimmune$pathway <- str_remove_all(string = trtGSEAimmune$pathway, pattern = "HALLMARK_")
trtGSEAimmune$pathway <- str_replace_all(string = trtGSEAimmune$pathway, pattern = "_", replacement = "\n")


ggplot(data = trtGSEAimmune,
       aes(x = pathway,
           y = trt,
           fill = biggerES)) +
  geom_tile() +
  geom_text(aes(label = round(biggerES, digits = 2))) +
  theme_minimal() + theme_bw() +
  scale_fill_gradientn(limits = c(-1,1),
                        colors = c(lowES,"white",highES))

ggsave(paste0(figPaperDir,"1d_trtGSEAES_female.png"), width = 8, height = 6)


#pval < .05 only
trtGSEAimmune <- trtGSEAimmune[trtGSEAimmune$pval_upreg < .05 | trtGSEAimmune$pval_downreg < .05,]

trtGSEAimmune$pathway <- ifelse(trtGSEAimmune$pathway == "IL6\nJAK\nSTAT3\nSIGNALING", "IL6 JAK STAT3\nSIGNALING", trtGSEAimmune$pathway)
trtGSEAimmune$pathway <- ifelse(trtGSEAimmune$pathway == "TNFA\nSIGNALING\nVIA\nNFKB", "TNFA\nSIGNALING VIA\nNFKB", trtGSEAimmune$pathway)


ggplot(data = trtGSEAimmune,
       aes(x = pathway,
           y = trt)) +
  geom_point(aes(col = biggerES, size = -log10(trtGSEAimmune$sigPval + 0.0001))) +
  #geom_text(aes(label = round(biggerES, digits = 2)), size = .8, col = "gray20") +
  theme_minimal() + theme_bw() +
  labs(size = "Pval") +
  scale_color_gradientn(limits = c(-1,1),
                        colors = c(lowES,"grey80",highES)) +
  theme(axis.text.y = element_text(size = 8, color = "black"),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 0.5, color = "black"),
        axis.title = element_blank(),
        legend.position = "none")
 ggsave(paste0(figPaperDir,"1dv2_trtGSEAES_female.png"), width = 4.5, height = 2.5)

  
ggplot(data = trtGSEAimmune,
       aes(x = pathway,
           y = trt)) +
  geom_point(aes(size = -log10(trtGSEAimmune$sigPval + 0.0001))) +
  #geom_text(aes(label = round(biggerES, digits = 2)), size = .8, col = "gray20") +
  theme_minimal() + theme_bw() +
  labs(size = "Pval") +
  scale_color_gradientn(limits = c(-1,1),
                        colors = c(lowES,"grey80",highES)) +
  theme(axis.text.y = element_text(size = 8, color = "black"),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 0.5, color = "black"),
        axis.title = element_blank())
 ggsave(paste0(figPaperDir,"1dv2_trtGSEAES_female_legend.png"), width = 4.5, height = 2.5)
```


```{r 1.E}
### 1E
# mean GE (resids) of genes in Inflammatory pathways, Ctrl v. Trt boxplot


for (hallPathName in c("HALLMARK_INFLAMMATORY_RESPONSE",
                       "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                       "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                       "HALLMARK_INTERFERON_GAMMA_RESPONSE")) {
  pathGeneList <- unique(subset(gene.set, GO.Term.Accession == hallPathName)$Associated.Gene.Name)
  
  for (sex in c("M","F","MF")) {
    if (sex == "M") {
      boxLines <- "gray80"
    } else {
      boxLines <- "gray10"
    }
    
    #trt<-"LPS"
    for (trt in c("LPS","Dex","Gard")) {
      
      modelName <- "G"
      if (trt == "LPS") {
        modelName <- "F"
      }
      
      residsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",trt,"_10000genes_5lCPM_model",modelName,"_resids.RDS"))
      residsPathF <- residsF[rownames(residsF) %in% pathGeneList,]
        
      metadataF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",trt,"_10000genes_5lCPM_metadata.RDS"))
      
      colnames(residsPathF) == metadataF$sampleID
      
      metadataF$pathMedian <- apply(residsPathF,2,median)
      
      if (trt == "LPS") {
        if (sex == "M") {
          trtColor <- lpsM
          ctrlColor <- ctrlM
        } else {
          trtColor <- lpsF
          ctrlColor <- ctrlF
        }
      } else if (trt == "Gard") {
        if (sex == "M") {
          trtColor <- gardM
          ctrlColor <- ctrlM
        } else {
          trtColor <- gardF
          ctrlColor <- ctrlF
        }
      } else {
        if (sex == "M") {
          trtColor <- dexM
          ctrlColor <- ctrlM
        } else {
          trtColor <- dexF
          ctrlColor <- ctrlF
        }
      }
      
      ggplot(data = metadataF,
             aes(x = Trt,
                 y = pathMedian,
                 fill = Trt)) +
        geom_boxplot(col = boxLines) +
        geom_hline(yintercept = 0, col = "gray30") +
        theme_minimal() + theme_bw() +
        #ylim(c(-2.75,2.75)) +
        ylab(paste0("Median GE ",hallPathName)) +
        scale_fill_manual(values = c(ctrlColor,trtColor))
      
      ggsave(paste0(figPaperDir,"1e_medPathwayGE_",sex,"_",hallPathName,"_",trt,".png"), width = 6, height = 6)
    }
  }
}


```

Figure 2
```{r}
#2.A
# plot of significant genes in M and F

#count of significant genes, male and female by treatment | using qvalue as a dummy
femaleLPS <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
#femaleLPS$qval_treatment <- qvalue(femaleLPS$pval_treatment)$qvalue

femaleGard <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Gardonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
#femaleGard$qval_treatment <- qvalue(femaleGard$pval_treatment)$qvalue

femaleDex <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Dexonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
#femaleDex$qval_treatment <- qvalue(femaleDex$pval_treatment)$qvalue

femaleCtrl <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
#femaleDex$qval_treatment <- qvalue(femaleDex$pval_treatment)$qvalue

femaleAll <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_allTrt_10000genes_5lCPM_modelG_allTrtF+Preg+WeightwKmatrix_emmreml")
#femaleDex$qval_treatment <- qvalue(femaleDex$pval_treatment)$qvalue


fdrThresh <- 0.10

trtSigGeneVector <- c(rep("mLPS",1),
                      rep("fLPS",sum(femaleLPS$fdr_status < fdrThresh)),
                      rep("mGard",1),
                      rep("fGard",sum(femaleGard$fdr_status < fdrThresh)),
                      rep("mDex",1),
                      rep("fDex",1),
                      rep("mCtrl",1),
                      rep("fCtrl",sum(femaleCtrl$fdr_status < fdrThresh)),
                      rep("mAll",1),
                      rep("fAll",sum(qvalue(femaleAll$pval_status)$qvalue < fdrThresh)))

trtSigGeneDF <- as.data.frame(trtSigGeneVector)

trtSigGeneDF$trtSigGeneVector <- factor(trtSigGeneDF$trtSigGeneVector, levels = c("fAll","mAll",
                                                                                  "fCtrl","mCtrl",
                                                                                  "fLPS","mLPS",
                                                                                  "fGard","mGard",
                                                                                  "fDex","mDex"))

length(unique(c(rownames(femaleLPS)[femaleLPS$fdr_status < fdrThresh],
         rownames(femaleCtrl)[femaleCtrl$fdr_status < fdrThresh],
         rownames(femaleGard)[femaleGard$fdr_status < fdrThresh])))


ggplot() +
  geom_bar(dat = trtSigGeneDF,
       aes(x = trtSigGeneVector,
           fill = trtSigGeneVector), width = 1/3) +
  geom_point(aes(x = "fAll",
                 y = sum(trtSigGeneDF$trtSigGeneVector == "fAll")), col = ctrlF, size = 12) +
  geom_point(aes(x = "fCtrl",
                 y = sum(trtSigGeneDF$trtSigGeneVector == "fCtrl")), col = ctrlF, size = 12) +
  geom_point(aes(x = "fLPS",
                 y = sum(trtSigGeneDF$trtSigGeneVector == "fLPS")), col = lpsF, size = 12) +
    geom_point(aes(x = "fGard",
                 y = sum(trtSigGeneDF$trtSigGeneVector == "fGard")), col = gardF, size = 10) +
  theme_minimal() + theme_bw() +
  scale_y_log10() +
  theme(axis.title = element_blank(),
        axis.text.y = element_text(size = 20),
        axis.text.x = element_text(size = 10)) +
  scale_fill_manual(values = c(ctrlF,ctrlM,ctrlF,ctrlM,lpsF,lpsM,gardF,gardM,dexF,dexM)) +
  #scale_color_manual(values = c(dexF,gardF,lpsF,dexM,gardM,lpsM)) +
  theme(legend.position = "none")

ggsave(paste0(figPaperDir,"2a_barplot.png"), width = 6, height = 3)


#data for venn diagram

lpsGenes <- rownames(femaleLPS)[femaleLPS$fdr_status < fdrThresh]
gardGenes <- rownames(femaleGard)[femaleGard$fdr_status < fdrThresh]
ctrlGenes <- rownames(femaleCtrl)[femaleCtrl$fdr_status < fdrThresh]
allGenes <- rownames(femaleAll)[qvalue(femaleAll$pval_status)$qvalue < fdrThresh]

(sqrt(length(lpsGenes))/pi) * 30
(sqrt(length(gardGenes))/pi) * 30
(sqrt(length(ctrlGenes))/pi) * 30
(sqrt(length(allGenes))/pi) * 30


#use an upset plot
###upset plot
listInput <- list(lpsGenes,
                  gardGenes,
                  ctrlGenes,
                  allGenes)

names(listInput) <- c("LPS",
                      "Gard",
                      "Ctrl",
                      "AllTrt")

testList <- fromList(listInput)

png(paste0(figPaperDir,"2a_upsetPlot_Status_fdr",fdrThresh,"_forVennDiagram.png"), width = 800, height = 800)
upset(data = testList, order.by = "freq", nsets = length(names(listInput)), nintersects = 20, text.scale = 2.5)
dev.off()

### plot variables
pointSize <- .75
axisSize <- .25
labelSize <- 20
numberSize <- 14

### gene list
allSigGenes <- unique(c(lpsGenes,gardGenes,ctrlGenes,allGenes))


## x-y plots
femaleAll$gene <- rownames(femaleAll)
femaleAll$stdBeta_status <- femaleAll$beta_status / sqrt(femaleAll$var_beta_status)

#ctrl - all
femaleCtrl$gene <- rownames(femaleCtrl)
femaleCtrl$stdBeta_status <- femaleCtrl$beta_status / sqrt(femaleCtrl$var_beta_status)
fAllCtrl <- merge(femaleAll,femaleCtrl, by = "gene", suffixes = c(".all",".ctrl"))

#fAllCtrlPlot <- fAllCtrl[fAllCtrl$fdr_status < fdrThresh | qvalue(fAllCtrl$pval_status.all)$qvalue < fdrThresh,]
fAllCtrlPlot <- fAllCtrl[fAllCtrl$gene %in% allSigGenes,]

#Gard
gardF <- "#E0B13F"

#LPS
lpsF <- "#E06E55"

#Dex
dexF <- "#4BC4E0"




ggplot(data = fAllCtrlPlot,
       aes(x = stdBeta_status.all,
           y = stdBeta_status.ctrl)) +
  geom_hline(yintercept = 0, lwd = axisSize) + geom_vline(xintercept = 0, lwd = axisSize) +
  geom_abline(slope = 1, intercept = 0, lwd = .5, linetype = "dashed") +
  theme_minimal() + theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
  xlim(c(-8,8)) + ylim(c(-8,8)) +
  scale_x_continuous(name="All Treatment", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  scale_y_continuous(name="Ctrl Condition", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  geom_point(size = pointSize, col = "gray15")
ggsave(paste0(figPaperDir,"3a_CtrlVallTrt_sigBetaPlot.png"), width = 4, height = 4)


#gard - all
femaleGard$gene <- rownames(femaleGard)
femaleGard$stdBeta_status <- femaleGard$beta_status / sqrt(femaleGard$var_beta_status)
fAllGard <- merge(femaleAll,femaleGard, by = "gene", suffixes = c(".all",".gard"))

#fAllGardPlot <- fAllGard[fAllGard$fdr_status < fdrThresh | qvalue(fAllGard$pval_status.all)$qvalue < fdrThresh,]
fAllGardPlot <- fAllGard[fAllGard$gene %in% allSigGenes,]

ggplot(data = fAllGardPlot,
       aes(x = stdBeta_status.all,
           y = stdBeta_status.gard)) +
  geom_hline(yintercept = 0, lwd = axisSize) + geom_vline(xintercept = 0, lwd = axisSize) +
  geom_abline(slope = 1, intercept = 0, lwd = .5, linetype = "dashed") +
  theme_minimal() + theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
  xlim(c(-8,8)) + ylim(c(-8,8)) +
  scale_x_continuous(name="All Treatment", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  scale_y_continuous(name="Gard Condition", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  geom_point(size = pointSize, col = gardF)
ggsave(paste0(figPaperDir,"3a_GardVallTrt_sigBetaPlot.png"), width = 4, height = 4)

ggplot(data = fAllGardPlot,
       aes(x = stdBeta_status.all,
           y = stdBeta_status.gard)) +
  geom_hline(yintercept = 0, lwd = axisSize) + geom_vline(xintercept = 0, lwd = axisSize) +
  geom_abline(slope = 1, intercept = 0, lwd = .5, linetype = "dashed") +
  theme_minimal() + theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
  xlim(c(-8,8)) + ylim(c(-8,8)) +
  scale_x_continuous(name="All Treatment", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  scale_y_continuous(name="Gard Condition", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5), position = "right") +
  geom_point(size = pointSize, col = gardF)
ggsave(paste0(figPaperDir,"3a_GardVallTrt_sigBetaPlot__RIGHT.png"), width = 4, height = 4)


#lps - all
#geneHighlight <- c("EDN1","SEMA4D","CD74","COMMD1","COMMD5","AKT1")
geneHighlight <- c("EDN1","COMMD1","COMMD5","AKT1")

femaleLPS$gene <- rownames(femaleLPS)
femaleLPS$stdBeta_status <- femaleLPS$beta_status / sqrt(femaleLPS$var_beta_status)
fAllLPS <- merge(femaleAll,femaleLPS, by = "gene", suffixes = c(".all",".lps"))

#fAllLPSPlot <- fAllLPS[fAllLPS$fdr_status < fdrThresh | qvalue(fAllLPS$pval_status.all)$qvalue < fdrThresh,]
fAllLPSPlot <- fAllLPS[fAllLPS$gene %in% allSigGenes,]

fAllLPSPlot$geneHighlight <- fAllLPSPlot$gene %in% geneHighlight

fAllLPSPlot <- fAllLPSPlot[order(fAllLPSPlot$geneHighlight),]

ggplot(data = fAllLPSPlot,
       aes(x = stdBeta_status.all,
           y = stdBeta_status.lps,
           col= geneHighlight)) +
  geom_hline(yintercept = 0, lwd = axisSize) + geom_vline(xintercept = 0, lwd = axisSize) +
  geom_abline(slope = 1, intercept = 0, lwd = .5, linetype = "dashed") +
  theme_minimal() + theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
  xlim(c(-8,8)) + ylim(c(-8,8)) +
  scale_x_continuous(name="All Treatment", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  scale_y_continuous(name="LPS Condition", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  geom_point(size = pointSize) +
  theme(legend.position = "none") +
  scale_color_manual(values = c(lpsF,"gray15"))
ggsave(paste0(figPaperDir,"3a_lpsVallTrt_sigBetaPlot.png"), width = 4, height = 4)


#dex - all
femaleDex$gene <- rownames(femaleDex)
femaleDex$stdBeta_status <- femaleDex$beta_status / sqrt(femaleDex$var_beta_status)
fAllDex <- merge(femaleAll,femaleDex, by = "gene", suffixes = c(".all",".dex"))

#fAllDexPlot <- fAllDex[fAllDex$fdr_status < fdrThresh | qvalue(fAllDex$pval_status.all)$qvalue < fdrThresh,]
fAllDexPlot <- fAllDex[fAllDex$gene %in% allSigGenes,]

ggplot(data = fAllDexPlot,
       aes(x = stdBeta_status.all,
           y = stdBeta_status.dex)) +
  geom_hline(yintercept = 0, lwd = axisSize) + geom_vline(xintercept = 0, lwd = axisSize) +
  geom_abline(slope = 1, intercept = 0, lwd = .5, linetype = "dashed") +
  theme_minimal() + theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
  xlim(c(-8,8)) + ylim(c(-8,8)) +
  scale_x_continuous(name="All Treatment", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  scale_y_continuous(name="Dex Condition", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  geom_point(size = pointSize, col = dexF)
ggsave(paste0(figPaperDir,"3a_DexVallTrt_sigBetaPlot.png"), width = 4, height = 4)

ggplot(data = fAllDexPlot,
       aes(x = stdBeta_status.all,
           y = stdBeta_status.dex)) +
  geom_hline(yintercept = 0, lwd = axisSize) + geom_vline(xintercept = 0, lwd = axisSize) +
  geom_abline(slope = 1, intercept = 0, lwd = .5, linetype = "dashed") +
  theme_minimal() + theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
  xlim(c(-8,8)) + ylim(c(-8,8)) +
  scale_x_continuous(name="All Treatment", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5)) +
  scale_y_continuous(name="Dex Condition", limits=c(-7.5, 7.5), breaks=c(-5,-2.5,0,2.5,5), position = "right") +
  geom_point(size = pointSize, col = dexF)
ggsave(paste0(figPaperDir,"3a_DexVallTrt_sigBetaPlot__RIGHT.png"), width = 4, height = 4)


```

```{r 3b single gene broken down by status and treatment}

ctrlM <- "grey40"
ctrlF <- "grey70"


geneList <- c("EDN1","IL16","CD74","IL6R","SEMA4D","GNG2","AKT1","COMMD1","COMMD5")

geneName <- "EDN1"

for (geneName in geneList) {
  pathGeneList <- geneName
  
  sex <- "F"
  trt <- "LPS"
  modelName <- "F"
  
  residsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",
                            trt,"_10000genes_5lCPM_model",modelName,"_resids.RDS"))
  
  residsPathF <- residsF[rownames(residsF) %in% pathGeneList,]
  
  metadataF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",
                              trt,"_10000genes_5lCPM_metadata.RDS"))
  
  emremlF <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",trt,
                            "_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_treatment"))
  emremlF$qval_status <- qvalue(emremlF$pval_status)$qvalue
  
  subset(emremlF, qval_status < .05 & fdr_treatment < .01)
  
  colnames(residsPathF) == metadataF$sampleID
      
  metadataF$pathMedian <- residsPathF
  metadataF$status_dom_sub <- factor(metadataF$status_dom_sub, levels = c("S","D"))
  
  metadataF$statTrt <- paste0(metadataF$status_dom_sub," ",metadataF$Trt)
  metadataF$statTrt <- factor(metadataF$statTrt, levels = c("S Ctrl","D Ctrl","S LPS","D LPS"))
  
  
  ### generate a list of paired Ctrl/LPS
  captPairs <- names(table(metadataF$captureref)[table(metadataF$captureref) > 1])
  metadataPairs <- subset(metadataF, captureref %in% captPairs)
  
  metadataPairs <- metadataPairs[order(metadataPairs$Trt),]
  metadataPairs <- metadataPairs[order(metadataPairs$captureref),]
  
  
  ggplot() +
    geom_boxplot(data = metadataF,
             aes(x = Trt,
                 y = pathMedian,
                 fill = statTrt)) +
    geom_hline(yintercept = 0) +
    scale_fill_manual(values = c(ctrlF,ctrlM,lpsM,lpsF)) +
        theme_minimal() + theme_bw() +
    theme(legend.position = "none") +
          theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = 24, color = "black"),
        axis.text = element_text(size = 24, color = "black"),
        axis.text.x = element_text(margin = margin(t = 30))) +
        #ylab(paste0(geneName," - Gene Expression")) +
    ggtitle(geneName) +
    xlab("") + ylab("Gene Expression")
  ggsave(paste0(figPaperDir,"3b_",sex,"_",geneName,"_",trt,".png"), width = 3, height = 6)
  
  ### gene expression response
  # "~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_modelF_responseResids.RDS"
  respResidsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,
                                "_",trt,"_10000genes_5lCPM_model",modelName,"_responseResids.RDS"))
  
  residsPathF <- respResidsF[rownames(respResidsF) %in% pathGeneList,]
  
  metadataF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",
                              trt,"_10000genes_5lCPM_responseMetadata.RDS"))
  
  colnames(residsPathF) == metadataF$sampleID
      
  metadataF$pathMedian <- residsPathF
  metadataF$status_dom_sub <- factor(metadataF$status_dom_sub, levels = c("S","D"))
  
  ggplot(data = metadataF,
             aes(x = status_dom_sub,
                 y = pathMedian,
                 fill = status_dom_sub)) +
    geom_hline(yintercept = 0) +
    geom_boxplot() +
    geom_jitter(height = 0, width = .05) +
    scale_fill_manual(values = c(lpsM,lpsF)) +
        theme_minimal() + theme_bw() +
    theme(legend.position = "none") +
        #ylab(paste0(geneName," - LogFC ",trt," Response")) +
      theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = 24, color = "black"),
        axis.text = element_text(size = 24, color = "black")) +
    ggtitle(geneName) +
    xlab("") + ylab("LPS Response")
  ggsave(paste0(figPaperDir,"3b_",sex,"_",geneName,"_",trt,"_LogFCResponse.png"), width = 2, height = 6)

  
  
  
}


```

```{r 3c}
geneList <- c( "HALLMARK_INTERFERON_ALPHA_RESPONSE",
               "HALLMARK_INTERFERON_GAMMA_RESPONSE",
               "HALLMARK_COMPLEMENT",
               "HALLMARK_IL6_JAK_STAT3_SIGNALING",
               "HALLMARK_INFLAMMATORY_RESPONSE",
               "HALLMARK_TNFA_SIGNALING_VIA_NFKB")

geneName <- "HALLMARK_INFLAMMATORY_RESPONSE"

for (geneName in geneList) {
  
  pathGeneList <- unique(subset(gene.set, GO.Term.Accession == geneName)$Associated.Gene.Name)
  
  sex <- "F"
  trt <- "LPS"
  modelName <- "F"
  
  residsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",
                            trt,"_10000genes_5lCPM_model",modelName,"_resids.RDS"))
  
  residsPathF <- residsF[rownames(residsF) %in% pathGeneList,]
  
  metadataF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",
                              trt,"_10000genes_5lCPM_metadata.RDS"))
  
  emremlF <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",trt,
                            "_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_treatment"))
  emremlF$qval_status <- qvalue(emremlF$pval_status)$qvalue
  
  subset(emremlF, qval_status < .05 & fdr_treatment < .01)
  
  colnames(residsPathF) == metadataF$sampleID
      
  metadataF$pathMedian <- apply(residsPathF,2,median)
  metadataF$pathMean <- apply(residsPathF,2,mean)
  metadataF$status_dom_sub <- factor(metadataF$status_dom_sub, levels = c("S","D"))
  
  ggplot(data = metadataF,
             aes(x = Trt,
                 y = pathMedian,
                 fill = status_dom_sub)) +
    geom_boxplot() +
        theme_minimal() + theme_bw() +
        ylab(paste0(geneName," - Gene Expression"))
  ggsave(paste0(figPaperDir,"3c_",sex,"_pathMedian",geneName,"_",trt,".png"), width = 6, height = 6)
  
  ggplot(data = metadataF,
             aes(x = Trt,
                 y = pathMedian,
                 fill = status_dom_sub)) +
    geom_boxplot() +
        theme_minimal() + theme_bw() +
        ylab(paste0(geneName," - Gene Expression"))
  ggsave(paste0(figPaperDir,"3c_",sex,"_pathMean",geneName,"_",trt,".png"), width = 6, height = 6)

}

```


```{r}
#sex * status * pathway median GE
### 2A
# mean GE (resids) of genes in Inflammatory pathways, Ctrl v. Trt boxplot

hallPathName <- "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
sex <- "F"
trt <- "LPS"

for (hallPathName in c("HALLMARK_INFLAMMATORY_RESPONSE",
                       "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                       "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                       "HALLMARK_INTERFERON_GAMMA_RESPONSE")) {
  pathGeneList <- unique(subset(gene.set, GO.Term.Accession == hallPathName)$Associated.Gene.Name)
  
  for (sex in c("F","M")) {
    
    if (sex == "M") {
      boxLines <- "gray80"
    } else {
      boxLines <- "gray10"
    }
    
    #trt<-"LPS"
    for (trt in c("LPS","Dex","Gard")) {
      
      modelName <- "G"
      if (trt == "LPS") {
        modelName <- "F"
      }
      
      residsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",trt,"only_10000genes_5lCPM_model",modelName,"_resids.RDS"))
      residsPathF <- residsF[rownames(residsF) %in% pathGeneList,]
        
      metadataF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",trt,"only_10000genes_5lCPM_metadata.RDS"))
      
      colnames(residsPathF) == metadataF$sampleID
      
      metadataF$pathMedian <- apply(residsPathF,2,median)
      
      if (trt == "LPS") {
        if (sex == "M") {
          trtColor <- lpsM
          ctrlColor <- ctrlM
        } else {
          trtColor <- lpsF
          ctrlColor <- ctrlF
        }
      } else if (trt == "Gard") {
        if (sex == "M") {
          trtColor <- gardM
          ctrlColor <- ctrlM
        } else {
          trtColor <- gardF
          ctrlColor <- ctrlF
        }
      } else {
        if (sex == "M") {
          trtColor <- dexM
          ctrlColor <- ctrlM
        } else {
          trtColor <- dexF
          ctrlColor <- ctrlF
        }
      }
      
      #metadataF$StatTrt <- paste0(metadataF$Trt,"-",metadataF$status_dom_sub)
      
      metadataF$status <- factor(metadataF$status_dom_sub, levels = c("S","D"))
      
      ggplot(data = metadataF,
             aes(x = status,
                 y = pathMedian,
                 fill = status)) +
        geom_violin(col = boxLines, draw_quantiles = c(.25,.5,.75)) +
        theme_minimal() + theme_bw() +
        #ylim(c(-2.75,2.75)) +
        ylab(paste0("Median GE ",hallPathName))
        #cale_alpha_manual(values = c(ctrlColor,trtColor))
      
      ggsave(paste0(figPaperDir,"2a_medPathwayGE_",sex,"_",hallPathName,"_",trt,".png"), width = 6, height = 6)
    }
  }
}

hallPathName <- "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
sex <- "F"
trt <- "LPS"

for (hallPathName in c("HALLMARK_INFLAMMATORY_RESPONSE",
                       "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                       "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                       "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                       "HALLMARK_COMPLEMENT",
                       "HALLMARK_CTRA_UP")) {
  pathGeneList <- unique(subset(gene.set, GO.Term.Accession == hallPathName)$Associated.Gene.Name)
  pathGenes <- unique(subset(gene.set, GO.Term.Accession == hallPathName)$Associated.Gene.Name)
    
    if (sex == "M") {
      boxLines <- "gray80"
    } else {
      boxLines <- "gray10"
    }
    
    #trt<-"LPS"
    ### WHY AM I LOOKING AT TRT-CTRL HERE THE SIGNAL IS IN ONLY not TRT CTRL
    for (trt in c("LPS","Dex","Gard","Ctrl")) {
      
      modelName <- "G"
      if (trt %in% c("LPS","Ctrl")) {
        modelName <- "F"
      }
      
      countsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_",trt,"only_10000genes_5lCPM_filteredCounts.RDS"))
      
      residsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_",trt,"only_10000genes_5lCPM_model",modelName,"_resids.RDS"))
      residsPathF <- residsF[rownames(residsF) %in% pathGeneList,]
      
      countsM <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_",trt,"only_10000genes_5lCPM_filteredCounts.RDS"))
      
      residsM <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_",trt,"only_10000genes_5lCPM_model",modelName,"_resids.RDS"))
      residsPathM <- residsM[rownames(residsM) %in% pathGeneList,]
      
      
      
      #get FDR sig genes
      trtEmrml <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_",trt,"only_10000genes_5lCPM_model",modelName,"_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status"))
      #ctrlEmrml <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status"))
      #sigGenes <- c(rownames(subset(trtEmrml, fdr_status < .2)),rownames(subset(ctrlEmrml, fdr_status < .2)))
      sigGenes <- rownames(subset(trtEmrml, fdr_status < .2))
      
      #top 25%
      pathGenesF <- rownames(residsPathF)
      pathCountsF <- countsF[rownames(countsF) %in% pathGenes,]
      
      pathCountsF$aveLogCPM <- aveLogCPM(pathCountsF)
      topGEpathGeneListF <- rownames(pathCountsF)[pathCountsF$aveLogCPM > summary(pathCountsF$aveLogCPM)[5]]
      residsPathTopGEF <- residsF[rownames(residsF) %in% topGEpathGeneListF,]
      
      pathGenesM <- rownames(residsPathM)
      pathCountsM <- countsM[rownames(countsM) %in% pathGenes,]
      
      pathCountsM$aveLogCPM <- aveLogCPM(pathCountsM)
      topGEpathGeneListM <- rownames(pathCountsM)[pathCountsM$aveLogCPM > summary(pathCountsM$aveLogCPM)[5]]
      residsPathTopGEM <- residsM[rownames(residsM) %in% topGEpathGeneListM,]

      
      #FDR < 0.20
      sigPathGenes <- intersect(sigGenes,pathGenes)
      
      residsPathSigGenesF <- residsF[rownames(residsF) %in% sigPathGenes,]
      residsPathSigGenesM <- residsM[rownames(residsM) %in% sigPathGenes,]
      
      metadataF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_",trt,"only_10000genes_5lCPM_metadata.RDS"))
      metadataM <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_",trt,"only_10000genes_5lCPM_metadata.RDS"))
      
      colnames(residsPathF) == metadataF$sampleID
      colnames(residsPathM) == metadataM$sampleID
      
      metadataF$pathMedian <- apply(residsPathF,2,median)
      metadataF$pathMedianTopGE <- apply(residsPathTopGEF,2,median)
      metadataF$pathMedianSigGenes <- apply(residsPathSigGenesF,2,median)
      
      metadataM$pathMedian <- apply(residsPathM,2,median)
      metadataM$pathMedianTopGE <- apply(residsPathTopGEM,2,median)
      metadataM$pathMedianSigGenes <- apply(residsPathSigGenesM,2,median)
      
      if (trt == "LPS") {
        if (sex == "M") {
          trtColor <- lpsM
          ctrlColor <- ctrlM
        } else {
          trtColor <- lpsF
          ctrlColor <- ctrlF
        }
      } else if (trt == "Gard") {
        if (sex == "M") {
          trtColor <- gardM
          ctrlColor <- ctrlM
        } else {
          trtColor <- gardF
          ctrlColor <- ctrlF
        }
      } else {
        if (sex == "M") {
          trtColor <- dexM
          ctrlColor <- ctrlM
        } else {
          trtColor <- dexF
          ctrlColor <- ctrlF
        }
      }
      
      metadataAll <- rbind.data.frame(metadataF,metadataM)
      
      #metadataAll$StatTrtSex <- paste0(metadataAll$Trt,"-",metadataAll$status_dom_sub,"_",metadataAll$sex)
      metadataAll$StatSex <- paste0(metadataAll$status_dom_sub,"_",metadataAll$sex)
      metadataAll$StatSex <- factor(metadataAll$StatSex, levels = c("S_M","D_M","S_F","D_F"))
      
      # if (trt == "LPS") {
      #   metadataAll$StatTrtSex <- factor(metadataAll$StatTrtSex, levels = c("Ctrl-S_M","Ctrl-D_M","Ctrl-S_F","Ctrl-D_F",
      #                                                                 "LPS-S_M","LPS-D_M","LPS-S_F","LPS-D_F"))
      # } else if (trt == "Dex") {
      #   metadataAll$StatTrtSex <- factor(metadataAll$StatTrtSex, levels = c("Ctrl-S_M","Ctrl-D_M","Ctrl-S_F","Ctrl-D_F",
      #                                                                 "Dex-S_M","Dex-D_M","Dex-S_F","Dex-D_F"))
      # } else {
      #   metadataAll$StatTrtSex <- factor(metadataAll$StatTrtSex, levels = c("Ctrl-S_M","Ctrl-D_M","Ctrl-S_F","Ctrl-D_F",
      #                                                                 "Gard-S_M","Gard-D_M","Gard-S_F","Gard-D_F"))
      # }
      
      #drop weird negative resid "SSUR2735"
      ggplot(data = metadataAll,
             aes(x = StatSex,
                 y = pathMedian,
                 col = StatSex,
                 fill = StatSex)) +
        geom_violin(col = boxLines, draw_quantiles = c(.25,.5,.75)) +
        #geom_beeswarm(draw_quantiles = c(.25,.5,.75)) +
        #geom_point() +
        theme_bw() +
        ylim(c(-1,1)) +
        ylab(paste0("Median GE ",hallPathName," all"))
        #scale_alpha_manual(values = c(ctrlColor,trtColor))
      
      ggsave(paste0(figPaperDir,"2a_medPathwayGE_",hallPathName,"_",trt,".png"), width = 6, height = 6)
      
      ggplot(data = metadataAll,
             aes(x = StatSex,
                 y = pathMedianTopGE,
                 fill = StatSex)) +
        geom_violin(col = boxLines, draw_quantiles = c(.25,.5,.75)) +
        theme_bw() +
        ylim(c(-1,1)) +
        ylab(paste0("Median GE ",hallPathName," top25%ge"))
      ggsave(paste0(figPaperDir,"2a_medPathwayGE_top25percent_",hallPathName,"_",trt,".png"), width = 6, height = 6)
      
      ggplot(data = metadataAll,
             aes(x = StatSex,
                 y = pathMedianSigGenes,
                 fill = StatSex)) +
        geom_violin(col = boxLines, draw_quantiles = c(.25,.5,.75)) +
        theme_bw() +
        ylim(c(-1,1)) +
        ylab(paste0("Median GE ",hallPathName," sig status genes"))
      ggsave(paste0(figPaperDir,"2a_medPathwayGE_sigGenes_",hallPathName,"_",trt,".png"), width = 6, height = 6)
  }
}

### what if we show MALE and FEMALE fold change within pathways?


### 2b TLR
# example gene plots
# GroupII IL6
# Group I / II / III / IV
lpsEmremlTrt <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_treatment")
lpsEmremlStatus <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

for (grp in c("I","II","III","IV")) {
  grpDir <- paste0(figPaperDir,"Group",grp,"/")
  dir.create(grpDir)
  
  if (grp == "I") {
    grpGenes <- intersect(rownames(lpsEmremlStatus)[lpsEmremlStatus$fdr_status < .1 & lpsEmremlStatus$beta_status < 0],
            rownames(lpsEmremlTrt)[lpsEmremlTrt$fdr_treatment < .1 & lpsEmremlTrt$beta_treatment > 0])
  } else if (grp == "II") {
    grpGenes <- intersect(rownames(lpsEmremlStatus)[lpsEmremlStatus$fdr_status < .1 & lpsEmremlStatus$beta_status > 0],
            rownames(lpsEmremlTrt)[lpsEmremlTrt$fdr_treatment < .1 & lpsEmremlTrt$beta_treatment > 0])
  } else if (grp == "III") {
    grpGenes <- intersect(rownames(lpsEmremlStatus)[lpsEmremlStatus$fdr_status < .1 & lpsEmremlStatus$beta_status < 0],
            rownames(lpsEmremlTrt)[lpsEmremlTrt$fdr_treatment < .1 & lpsEmremlTrt$beta_treatment < 0])
  } else if (grp == "IV") {
    grpGenes <- intersect(rownames(lpsEmremlStatus)[lpsEmremlStatus$fdr_status < .1 & lpsEmremlStatus$beta_status > 0],
            rownames(lpsEmremlTrt)[lpsEmremlTrt$fdr_treatment < .1 & lpsEmremlTrt$beta_treatment < 0])
  }
  
  print(grp)
  print(length(grpGenes))
  
  #geneName <- "IL6"
  
    for (geneName in grpGenes) {
      residsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_modelF_resids.RDS"))
      metadataF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_metadata.RDS"))
      
      all(colnames(residsF) == metadataF$sampleID)
      
      metadataF$singleGene <- residsF[which(rownames(residsF) == geneName),]
      
      metadataF$TrtStatus <- paste0(metadataF$Trt,"-",metadataF$status_dom_sub)
      
      metadataF$status_dom_sub <- factor(metadataF$status_dom_sub, levels = c("S","D"))
      
      ggplot(data = metadataF,
             aes(x = status_dom_sub,
                 y = singleGene,
                 col = Trt)) +
        #geom_violin(draw_quantiles = c(.25,.5,.75)) +
        geom_jitter(height = 0, width = .1) +
        geom_smooth(aes(group = Trt),method = "lm", formula = "y~x") +
        theme_minimal() + theme_bw() + 
        xlab("Status") + ylab(geneName) +
        scale_color_manual(values = c(ctrlF,lpsF))
      
      ggsave(paste0(grpDir,"2b_GeneExample_",geneName,".png"), width = 4, height = 3)
      
      residsFresp <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_modelF_responseResids.RDS"))
      metadataFresp <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_responseMetadata.RDS"))
      
      metadataFresp$singleGene <- residsFresp[which(rownames(residsFresp) == geneName),]
      metadataFresp$status_dom_sub <- factor(metadataFresp$status_dom_sub, levels = c("S","D"))
      
      ggplot(data = metadataF,
             aes(x = status_dom_sub,
                 y = singleGene,
                 fill = status_dom_sub)) +
        geom_violin(aes(alpha = status_dom_sub), draw_quantiles = c(.25,.5,.75)) +
        geom_beeswarm(alpha = .5) +
        geom_hline(yintercept = 0) +
        #geom_smooth(aes(group = Trt),method = "lm", formula = "y~x") +
        theme_bw() + theme(legend.position = "none") +
        xlab("Status") + ylab(geneName) +
        scale_alpha_manual(values = c(.6,1)) +
        scale_fill_manual(values = c(lpsF,lpsF))
      ggsave(paste0(grpDir,"2b_GeneExample_Response_",geneName,".png"), width = 4, height = 3)
    }
}
```



Figure 3
```{r}

#3a - upset plot
softCutoff <- TRUE
#upset plot of shared status genes

FDRthresh <- 0.10
FDRSoftCut <- 0.10

#get list of FDR sig Status genes, separate up and down
femaleLPS <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

femaleGard <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Gardonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

femaleDex <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Dexonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

femaleCtrl <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

LPSup <- rownames(subset(femaleLPS, beta_status > 0 & fdr_status < FDRthresh))
LPSdown <- rownames(subset(femaleLPS, beta_status < 0 & fdr_status < FDRthresh))
Gardup <- rownames(subset(femaleGard, beta_status > 0 & fdr_status < FDRthresh))
Garddown <- rownames(subset(femaleGard, beta_status < 0 & fdr_status < FDRthresh))
Dexup <- rownames(subset(femaleDex, beta_status > 0 & fdr_status < FDRthresh))
Dexdown <- rownames(subset(femaleDex, beta_status < 0 & fdr_status < FDRthresh))
Ctrlup <- rownames(subset(femaleCtrl, beta_status > 0 & fdr_status < FDRthresh))
Ctrldown <- rownames(subset(femaleCtrl, beta_status < 0 & fdr_status < FDRthresh))

if (softCutoff) {
LPSupSoft <- rownames(subset(femaleLPS, beta_status > 0 & fdr_status < (FDRthresh + FDRSoftCut) & fdr_status > FDRthresh))
LPSdownSoft <- rownames(subset(femaleLPS, beta_status < 0 & fdr_status < (FDRthresh + FDRSoftCut) & fdr_status > FDRthresh))
GardupSoft <- rownames(subset(femaleGard, beta_status > 0 & fdr_status < (FDRthresh + FDRSoftCut) & fdr_status > FDRthresh))
GarddownSoft <- rownames(subset(femaleGard, beta_status < 0 & fdr_status < (FDRthresh + FDRSoftCut) & fdr_status > FDRthresh))
DexupSoft <- rownames(subset(femaleDex, beta_status > 0 & fdr_status < (FDRthresh + FDRSoftCut) & fdr_status > FDRthresh))
DexdownSoft <- rownames(subset(femaleDex, beta_status < 0 & fdr_status < (FDRthresh + FDRSoftCut) & fdr_status > FDRthresh))
CtrlupSoft <- rownames(subset(femaleCtrl, beta_status > 0 & fdr_status < (FDRthresh + FDRSoftCut) & fdr_status > FDRthresh))
CtrldownSoft <- rownames(subset(femaleCtrl, beta_status < 0 & fdr_status < (FDRthresh + FDRSoftCut) & fdr_status > FDRthresh))

LPSup <- c(LPSup,LPSupSoft[LPSupSoft %in% c(Gardup,Garddown,Dexup,Dexdown,Ctrlup,Ctrldown)])
LPSdown <- c(LPSdown,LPSdownSoft[LPSdownSoft %in% c(Gardup,Garddown,Dexup,Dexdown,Ctrlup,Ctrldown)])

Gardup <- c(Gardup,GardupSoft[GardupSoft %in% c(LPSup,LPSdown,Dexup,Dexdown,Ctrlup,Ctrldown)])
Garddown <- c(Garddown,GarddownSoft[GarddownSoft %in% c(LPSup,LPSdown,Dexup,Dexdown,Ctrlup,Ctrldown)])

Dexup <- c(Dexup,DexupSoft[DexupSoft %in% c(Gardup,Garddown,LPSup,LPSdown,Ctrlup,Ctrldown)])
Dexdown <- c(Dexdown,DexdownSoft[DexdownSoft %in% c(Gardup,Garddown,LPSup,LPSdown,Ctrlup,Ctrldown)])

Ctrlup <- c(Ctrlup,CtrlupSoft[CtrlupSoft %in% c(Gardup,Garddown,Dexup,Dexdown,LPSup,LPSdown)])
Ctrldown <- c(Ctrldown,CtrldownSoft[CtrldownSoft %in% c(Gardup,Garddown,Dexup,Dexdown,LPSup,LPSdown)])
}

###upset plot
listInput <- list(LPSup,
                  LPSdown,
                  Gardup,
                  Garddown,
                  Dexup,
                  Dexdown,
                  Ctrlup,
                  Ctrldown)

names(listInput) <- c("LPSup",
                  "LPSdown",
                  "Gardup",
                  "Garddown",
                  "Dexup",
                  "Dexdown",
                  "Ctrlup",
                  "Ctrldown")

testList <- fromList(listInput)

if (softCutoff) {
  png(paste0(figPaperDir,"3a_upsetPlot_Status_fdr",FDRthresh,"_soft",FDRSoftCut,".png"), width = 800, height = 800)
} else {
  png(paste0(figPaperDir,"3a_upsetPlot_Status_fdr",FDRthresh,".png"), width = 800, height = 800)
}
upset(data = testList, order.by = "freq", nsets = length(names(listInput)), nintersects = 20, text.scale = 2.5)
dev.off()

#3b 
# Dex beta-beta plot
femaleLPS <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
femaleGard <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Gardonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
femaleDex <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Dexonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
femaleCtrl <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

#beta-beta by pathway
#inflamm
c("HALLMARK_INFLAMMATORY_RESPONSE","HALLMARK_TNFA_SIGNALING_VIA_NFKB")
#interferon
c("HALLMARK_INTERFERON_ALPHA_RESPONSE","HALLMARK_INTERFERON_GAMMA_RESPONSE")

trtX <- "LPS"
trtY <- "Gard"

for (trtX in c("LPS","Gard","Ctrl","Dex")) {
  for (trtY in c("LPS","Gard","Ctrl","Dex")) {    
    emrmlX <- get(paste0("female",trtX))
    emrmlY <- get(paste0("female",trtY))
    
    pathGeneList <- unique(subset(gene.set, GO.Term.Accession %in% c("HALLMARK_INFLAMMATORY_RESPONSE","HALLMARK_TNFA_SIGNALING_VIA_NFKB"))$Associated.Gene.Name)
    
    #common set
    commonGenes <- unique(intersect(intersect(rownames(emrmlX),rownames(emrmlY)),pathGeneList))
    
    compTrtX <- emrmlX[rownames(emrmlX) %in% commonGenes,]; compTrtY <- emrmlY[rownames(emrmlY) %in% commonGenes,]
    
    compTrtX$gene <- rownames(compTrtX); compTrtY$gene <- rownames(compTrtY)
    
    compBetas <- merge(compTrtX,compTrtY, by = "gene", suffixes = c(".x",".y"))
    compR <- round(cor(compBetas$beta_status.x,compBetas$beta_status.y), digits = 2)
    
    ggplot(compBetas,
           aes(x = beta_status.x,
               y = beta_status.y)) +
      geom_point() +
      geom_smooth(method = "lm", formula = "y~x") +
      xlab(paste0("Status Beta ",trtX," only")) +
      ylab(paste0("Status Beta ",trtY," only")) +
      ggtitle(paste0(trtY," v. ",trtX," Inflammatory Gene Set Betas"),
              subtitle = paste0("R=",compR))
    ggsave(paste0(figPaperDir,"3c_beta-beta_Inflamm_pathways_",trtY,"v",trtX,".png"), width = 4, height = 4)
  }
}


for (trtX in c("LPS","Gard","Ctrl","Dex")) {
  for (trtY in c("LPS","Gard","Ctrl","Dex")) {    
    emrmlX <- get(paste0("female",trtX))
    emrmlY <- get(paste0("female",trtY))
    
    pathGeneList <- unique(subset(gene.set, GO.Term.Accession %in% c("HALLMARK_INTERFERON_ALPHA_RESPONSE","HALLMARK_INTERFERON_GAMMA_RESPONSE"))$Associated.Gene.Name)
    
    #common set
    commonGenes <- unique(intersect(intersect(rownames(emrmlX),rownames(emrmlY)),pathGeneList))
    
    compTrtX <- emrmlX[rownames(emrmlX) %in% commonGenes,]; compTrtY <- emrmlY[rownames(emrmlY) %in% commonGenes,]
    
    compTrtX$gene <- rownames(compTrtX); compTrtY$gene <- rownames(compTrtY)
    
    compBetas <- merge(compTrtX,compTrtY, by = "gene", suffixes = c(".x",".y"))
    compR <- round(cor(compBetas$beta_status.x,compBetas$beta_status.y), digits = 2)
    
    ggplot(compBetas,
           aes(x = beta_status.x,
               y = beta_status.y)) +
      geom_point() +
      geom_smooth(method = "lm", formula = "y~x") +
      xlab(paste0("Status Beta ",trtX," only")) +
      ylab(paste0("Status Beta ",trtY," only")) +
      ggtitle(paste0(trtY," v. ",trtX," Interferon Gene Set Betas"),
              subtitle = paste0("R=",compR))
    ggsave(paste0(figPaperDir,"3c_beta-beta_Interferon_pathways_",trtY,"v",trtX,".png"), width = 4, height = 4)
  }
}
```

Figure 4 Resampling 
```{r}
geneSet <- unique(gene.set)
geneSet <- geneSet[order(geneSet$Category),]

#X-Y plot 
## X = Population Beta
## Y = Resampling T-Statistic
longSampleResults <- readRDS(file = paste0(outDir,"longitudinalSampleResults.RDS"))
longSampleResultsCtrl <- readRDS(file = paste0(outDir,"longitudinalSampleResultsCtrl.RDS"))



longSampleResultsPlot <- subset(longSampleResults, emmremlQValStatus < .1)
longSampleResultsCtrlPlot <- subset(longSampleResultsCtrl, emmremlQValStatus < .1)

longSampleResultsCtrlPlot <- longSampleResultsCtrlPlot[longSampleResultsCtrlPlot$geneNamePlot %in% longSampleResultsPlot$geneNamePlot,]


yMax <- ceiling(max(abs(c(longSampleResultsPlot$tTestStatStatus,longSampleResultsCtrlPlot$tTestStatStatus))))
xMax <- ceiling(max(abs(c(longSampleResultsPlot$emmremlStdBetaStatus,longSampleResultsCtrlPlot$emmremlStdBetaStatus))))


ggplot(data = longSampleResults,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  geom_point() + theme_minimal() + theme_bw() +
  ylab("Longitudinal Sample T-Test") +
  xlab("Population Sample Status Beta")


longSampleResults$immune <- FALSE
longSampleResults$signal <- FALSE

longSampleResults$IFNalpha <- FALSE
longSampleResults$IFNgamma <- FALSE
longSampleResults$TNFANFKB <- FALSE
longSampleResults$Inflamm <- FALSE
longSampleResults$Complement <- FALSE
longSampleResults$IL6JAK <- FALSE



r <- 2

for (r in 1:dim(longSampleResults)[1]) {
  if (longSampleResults$geneNamePlot[r] %in% geneSet$Associated.Gene.Name) {
    paths <- geneSet$GO.Term.Accession[which(geneSet$Associated.Gene.Name == longSampleResults$geneNamePlot[r])]
    cats <- geneSet$Category[which(geneSet$Associated.Gene.Name == longSampleResults$geneNamePlot[r])]
    
    #categories
    if ("immune" %in% cats) {
      longSampleResults$immune[r] <- TRUE
    }
    if ("signaling" %in% cats) {
      longSampleResults$signal[r] <- TRUE
    }
    
    #pathways
    if ("HALLMARK_INTERFERON_ALPHA_RESPONSE" %in% paths) {
      longSampleResults$IFNalpha[r] <- TRUE
    }
    if ("HALLMARK_INTERFERON_GAMMA_RESPONSE" %in% paths) {
      longSampleResults$IFNgamma[r] <- TRUE
    }
    if ("HALLMARK_TNFA_SIGNALING_VIA_NFKB" %in% paths) {
      longSampleResults$TNFANFKB[r] <- TRUE
    }
    if ("HALLMARK_INFLAMMATORY_RESPONSE" %in% paths) {
      longSampleResults$Inflamm[r] <- TRUE
    }
    if ("HALLMARK_COMPLEMENT" %in% paths) {
      longSampleResults$Complement[r] <- TRUE
    }
    if ("HALLMARK_IL6_JAK_STAT3_SIGNALING" %in% paths) {
      longSampleResults$IL6JAK[r] <- TRUE
    }
  }
}

longSampleResultsPlot <- longSampleResults[order(longSampleResults$immune, decreasing = FALSE),]

ggplot(data = longSampleResultsPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus,
           col = immune)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  geom_point() + theme_minimal() + theme_bw() +
  scale_color_manual(values = c("grey90","orange")) +
  ylab("Longitudinal Sample T-Test") +
  xlab("Population Sample Status Beta")

longSampleResults$immuneORsignal <- longSampleResults$immune | longSampleResults$signal

longSampleResultsPlot <- longSampleResults[order(longSampleResults$immuneORsignal, decreasing = FALSE),]

ggplot(data = longSampleResultsPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus,
           col = immuneORsignal)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  geom_point(size = 4) + theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 14)) +
  scale_color_manual(values = c("grey90","orange")) +
  ylab("Longitudinal Sample T-Test") +
  xlab("Population Sample Status Beta")
ggsave(paste0(figPaperDir,"4a_longResample_ttest_v_emreml_statusEffect_immuneORsignal.png"), width = 6, height = 4)

#longSampleResults$keyPathway <- longSampleResults$IFNalpha | longSampleResults$IFNgamma | longSampleResults$TNFANFKB | longSampleResults$Inflamm | longSampleResults$Complement | longSampleResults$IL6JAK

longSampleResults$keyPathway <- longSampleResults$IFNgamma | longSampleResults$TNFANFKB | longSampleResults$Inflamm | longSampleResults$Complement

longSampleResultsPlot <- longSampleResults[order(longSampleResults$keyPathway, decreasing = FALSE),]

ggplot(data = longSampleResultsPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus,
           col = keyPathway)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  geom_point(size = 4) + theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 14)) +
  scale_color_manual(values = c("grey90","coral")) +
  ylab("Longitudinal Sample T-Test") +
  xlab("Population Sample Status Beta")
ggsave(paste0(figPaperDir,"4a_longResample_ttest_v_emreml_statusEffect_keyPathway.png"), width = 6, height = 4)


longSampleResultsPlot <- longSampleResults[order(longSampleResults$Inflamm, decreasing = FALSE),]

ggplot(data = longSampleResultsPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus,
           col = Inflamm)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  geom_point(size = 4) + theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 14)) +
  scale_color_manual(values = c("grey90","red")) +
  ylab("Longitudinal Sample T-Test") +
  xlab("Population Sample Status Beta")
ggsave(paste0(figPaperDir,"4a_longResample_ttest_v_emreml_statusEffect_Inflammation.png"), width = 6, height = 4)

#with up/down colors
longSampleResultsPlot$keyPathUpDown <- ifelse(longSampleResultsPlot$tTestStatStatus > 0, "up","down")
longSampleResultsPlot$keyPathUpDown <- ifelse(longSampleResultsPlot$keyPathway, longSampleResultsPlot$keyPathUpDown, "none")

longSampleResultsPlot <- longSampleResultsPlot[order(longSampleResultsPlot$keyPathway, decreasing = FALSE),]

ggplot(data = longSampleResultsPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus,
           col = keyPathUpDown)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  geom_point(size = 4) + theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 14)) +
  scale_color_manual(values = c("#406F8B","grey90","#A3D65A")) +
  ylim(c(-yMax,yMax)) + xlim(c(-xMax,xMax)) +
  ylab("Longitudinal Sample T-Test") +
  xlab("Population Sample Status Beta")
ggsave(paste0(figPaperDir,"4a_longResample_ttest_v_emreml_statusEffect_keyPathwayUpDown.png"), width = 6, height = 4)



## 4B control long samples




plotSigOnly <- TRUE
if (plotSigOnly) {
  longSampleResultsCtrlPlot <- subset(longSampleResultsCtrl, emmremlQValPreg < .1)
} else {
  longSampleResultsCtrlPlot <- longSampleResultsCtrl
}

fdrThresh <- .2

ggplot(longSampleResultsCtrlPlot) +
geom_histogram(aes(x = tTestStatPreg,
                   fill = ifelse(emmremlQValPreg < fdrThresh,
                                 ifelse(emmremlStdBetaPreg > 0,"sigUp","sigDn"),"n.s.")), bins = 50) +
labs(fill = "sig in\nemmreml")

ggplot(longSampleResultsCtrlPlot) +
geom_histogram(aes(x = meanGEpreg - meanGEnonpreg,
                   fill = ifelse(emmremlQValPreg < fdrThresh,
                                 ifelse(emmremlStdBetaPreg > 0,"sigUp","sigDn"),"n.s.")), bins = 50) +
labs(fill = "sig in\nemmreml")

if (plotSigOnly) {
  longSampleResultsCtrlPlot <- subset(longSampleResultsCtrl, emmremlQValStatus < .1)
} else {
  longSampleResultsCtrlPlot <- longSampleResultsCtrlPlot
}


ggplot(longSampleResultsCtrlPlot) +
  geom_histogram(aes(x = tTestStatStatus,
                   fill = ifelse(emmremlQValStatus < fdrThresh,
                                 ifelse(emmremlStdBetaStatus > 0,"sigUp","sigDn"),"n.s.")), bins = 50) +
labs(fill = "sig in\nemmreml")

ggplot(longSampleResultsCtrlPlot) +
  geom_point(aes(x = emmremlStdBetaStatus,
                 y = tTestStatStatus,
                   col = wTestPValStatus < 0.05)) +
labs(col = "sig in\nT-Test")

### match color of test plot
longSampleResultsCtrlPlot$keyPathUpDown <- ifelse(longSampleResultsCtrlPlot$geneNamePlot %in% subset(longSampleResultsPlot, keyPathUpDown == "up")$geneNamePlot, "up",
                                                  ifelse(longSampleResultsCtrlPlot$geneNamePlot %in% subset(longSampleResultsPlot, keyPathUpDown == "down")$geneNamePlot, "down","zone"))
longSampleResultsCtrlPlot <- longSampleResultsCtrlPlot[order(longSampleResultsCtrlPlot$keyPathUpDown, decreasing = TRUE),]


ggplot(data = longSampleResultsCtrlPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  ylim(c(-11,11)) +
  geom_point(size = 2, col = "gray65") + theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = 16, color = "black"),
        axis.title = element_text(size = 14, color = "black"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ylim(c(-yMax,yMax)) + xlim(c(-xMax,xMax)) +
  ylab("Longitudinal Sample T-Test") +
  xlab("Population Sample Status Beta")
ggsave(paste0(figPaperDir,"4a_longResample_inSub_ttest_v_emreml_statusEffect.png"), width = 4, height = 4)

ggplot(data = longSampleResultsCtrlPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus,
           col = keyPathUpDown)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  ylim(c(-11,11)) +
  geom_point(size = 2) + theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = 16, color = "black"),
        axis.title = element_text(size = 14, color = "black"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c("#406F8B","#A3D65A","grey65")) +
  ylim(c(-yMax,yMax)) + xlim(c(-xMax,xMax)) +
  ylab("Longitudinal Sample T-Test") +
  xlab("Population Sample Status Beta")
ggsave(paste0(figPaperDir,"4a_longResample_inSub_ttest_v_emreml_statusEffect_keyPathwayUpDown.png"), width = 4, height = 4)

ggplot(data = longSampleResultsPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus,
           col = keyPathUpDown)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  geom_point(size = 2) + theme_minimal() + theme_bw() +
  theme(axis.text = element_text(size = 10, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = c("#406F8B","grey75","#A3D65A")) +
  ylim(c(-yMax,yMax)) + xlim(c(-xMax,xMax)) +
  ylab("Dominance Status Effect\n(T-statistic; Longitudinal Sample)") +
  xlab("Dominance Status Effect\n(Cross-Sectional Sample)")
ggsave(paste0(figPaperDir,"4a_longResample_ttest_v_emreml_statusEffect_keyPathwayUpDown.png"), width = 4, height = 4)



length(longSampleResultsCtrlPlot$statusDirConsistent)
binom.test(313,514,.5)

sum(longSampleResultsCtrl$statusSigConsistent[longSampleResultsCtrl$geneNamePlot %in% longSampleResultsPlot$geneNamePlot[longSampleResultsPlot$statusSigConsistent]])


prop.test(c(313,454),c(514,514),alternative = "less")$p.value
fisher.test(matrix(c(313,454,514,514), nrow = 2, byrow = T), alternative = "less")



testCor <- cor.test(longSampleResultsPlot$emmremlStdBetaStatus, longSampleResultsPlot$tTestStatStatus, method = "pearson")
testCor$estimate
testCor$p.value

ctrlCor <- cor.test(longSampleResultsCtrlPlot$emmremlStdBetaStatus, longSampleResultsCtrlPlot$tTestStatStatus, method = "pearson")
ctrlCor$estimate
ctrlCor$p.value

ggplot(data = longSampleResultsCtrlPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  ylim(c(-11,11)) +
  geom_point(size = 2, alpha = .2) + theme_minimal() + theme_bw() +
  geom_smooth(formula = "y~x", method = "lm")

ggplot(data = longSampleResultsPlot,
       aes(x = emmremlStdBetaStatus,
           y = tTestStatStatus)) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  ylim(c(-11,11)) +
  geom_point(size = 2, alpha = .2) + theme_minimal() + theme_bw() +
  geom_smooth(formula = "y~x", method = "lm")

```


```{r gene inset}
#genes that change in population and individuals across status change


geneList <- c("IL7R","SERPINB2","BANK1","SEMA4D")

### for each gene in geneList, plot the individal change in GE pre-post status change
geneName <- "BANK1"
for (geneName in geneList) {
  pathGeneList <- geneName
  
  sex <- "F"
  trt <- "LPS"
  modelName <- "F"
  
  residsF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",trt,"only_10000genes_5lCPM_model",modelName,"_resids.RDS"))
  
  residsPathF <- residsF[rownames(residsF) %in% pathGeneList,]
  
  metadataF <- readRDS(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_",sex,"_",trt,"only_10000genes_5lCPM_metadata.RDS"))
  
  colnames(residsPathF) == metadataF$sampleID
      
  metadataF$pathMedian <- residsPathF
  metadataF$status_dom_sub <- factor(metadataF$status_dom_sub, levels = c("S","D"))
  
  ggplot(data = metadataF,
             aes(x = status_dom_sub,
                 y = pathMedian)) +
    geom_boxplot(col = "gray70") +
    geom_jitter(width = .1, height = 0, col = "gray70") +
        #geom_hline(yintercept = 0, col = "gray30") +
        theme_minimal() + theme_bw() +
        ylab(paste0(geneName," - Gene Expression"))
  ggsave(paste0(figPaperDir,"4c_medGE_",sex,"_",geneName,"_",trt,".png"), width = 6, height = 6)
  
  #now highlight the resampled individuals
  #dom-dub resamp
    ### Find the repeat animals S/D
  repeatSamples <- names(table(metadataF$correctsampleID)[table(metadataF$correctsampleID) > 1])
  
  ### find the repeat animals preg/not-preg
  s <- repeatSamples[1]
  doubleStatus <- vector()
  sampleID <- vector()
  pregTF <- vector()
  status <- vector()
  animalID <- vector()
  
  for (s in repeatSamples) {
    #print(s)
    if (length(unique(subset(metaStatus, animal_id == s)$status_dom_sub)) > 1) {
      #print("duplicate")
      doubleStatus <- c(doubleStatus,s)
      
      animalMeta <- subset(metaStatus, animal_id == s)
      sampleID <- c(sampleID,animalMeta$sampleID)
      pregTF <- c(pregTF,animalMeta$pregStatusBackDate35)
      status <- c(status,animalMeta$status_dom_sub)
      animalID <- c(animalID,animalMeta$animal_id)
      
    } else {
      #print("nope")
    }
  }
  
  statusResample <- cbind.data.frame(sampleID,pregTF,status,animalID)
  
  pregMismatch <- c("SSUR1400","SSUR1485","SSUR1474","SSUR1605","SSUR1781","SSUR2229","SSUR1862","SSUR1905","SSUR1472")
  statusResample <- statusResample[!statusResample$sampleID %in% pregMismatch,]
  
  #pull data
  metadataF$resample <- ifelse(metadataF$correctsampleID %in% statusResample$animalID, TRUE, FALSE)
  metadataFResample <- subset(metadataF, resample)
  
  ## use a loop to calclate the median before and after for each individual
  females <- unique(metadataFResample$correctsampleID)
  S <- vector()
  D <- vector()
  
  f <- "VVHF110"
  for (f in females) {
    if (geneName == "SERPINB2") {
      S <- c(S,min(subset(metadataFResample, correctsampleID == f & status_dom_sub == "S")$pathMedian))
      D <- c(D,max(subset(metadataFResample, correctsampleID == f & status_dom_sub == "D")$pathMedian))
    } else {
      S <- c(S,max(subset(metadataFResample, correctsampleID == f & status_dom_sub == "S")$pathMedian))
      D <- c(D,min(subset(metadataFResample, correctsampleID == f & status_dom_sub == "D")$pathMedian))
    }
  }
  
  resamplePlot <- melt(cbind.data.frame(females,S,D))
  colnames(resamplePlot) <- c("females","status_dom_sub","pathMedian")
  
  #add to plot
  ggplot() +
    geom_boxplot(aes(x = metadataF$status_dom_sub,
                     y = metadataF$pathMedian), col = "gray70") +
    geom_point(aes(x = metadataF$status_dom_sub,
                     y = metadataF$pathMedian), col = "gray70") +
    geom_point(aes(x = resamplePlot$status_dom_sub,
                   y = resamplePlot$pathMedian)) +
    geom_line(aes(x = resamplePlot$status_dom_sub,
                   y = resamplePlot$pathMedian,
                  group = resamplePlot$females)) +
        #geom_hline(yintercept = 0, col = "gray30") +
        theme_minimal() + theme_bw() +
        ylab(paste0(geneName," - Gene Expression"))
  ggsave(paste0(figPaperDir,"4c_medGE_",sex,"_",geneName,"_",trt,"_withResample.png"), width = 6, height = 6)
}

```


```{r pathway inset}
#genes that change in population and individuals across status change

geneList <- longSampleResults$geneNamePlot[longSampleResults$keyPathway]

trt <- "LPS"

lpsPlot <- subset(longSampleResults, geneNamePlot %in% geneList & treatment == "LPS")

lpsPlot <- lpsPlot[,colnames(lpsPlot) %in% c("emmremlStdBetaStatus","geneNamePlot","meanGEdom","meanGEsub")]

colnames(lpsPlot)[1:3] <- c("geneName","D","S")

lpsPlotUp <- subset(lpsPlot, emmremlStdBetaStatus > 0)
lpsPlotDown <- subset(lpsPlot, emmremlStdBetaStatus < 0)

lpsPlotMelt <- melt(lpsPlot[,1:3])
lpsPlotUp <- melt(lpsPlotUp[,1:3])
lpsPlotDown <- melt(lpsPlotDown[,1:3])

lpsPlotMelt$variable <- factor(lpsPlotMelt$variable, levels = c("S","D"))
lpsPlotUp$variable <- factor(lpsPlotUp$variable, levels = c("S","D"))
lpsPlotDown$variable <- factor(lpsPlotDown$variable, levels = c("S","D"))

ggplot(data = lpsPlotUp,
       aes(x = variable,
           y = value,
           group = geneName)) +
  geom_line(linewidth = 1.5, col = "coral") +
  theme_minimal() + theme_bw() + 
  ylab("Mean GE in Long. Samples") + xlab("Status") +
  ggtitle("Key Pathway Genes Sig Up in Dom F Meerkats",
          subtitle = paste0("Resampled GE Values, ",trt,", Treatment, FDR < 0.10"))
ggsave(paste0(figPaperDir,"4b_Resample_GeneValues_keyPathsLPS_Up.png"), width = 4, height = 4)

ggplot(data = lpsPlotDown,
       aes(x = variable,
           y = value,
           group = geneName)) +
  geom_line(linewidth = 1.5, col = "coral") +
  theme_minimal() + theme_bw() + 
  ylab("Mean GE in Long. Samples") + xlab("Status") +
  ggtitle("Key Pathway Genes Sig Down in Dom F Meerkats",
          subtitle = paste0("Resampled GE Values, ",trt,", Treatment, FDR < 0.10"))
ggsave(paste0(figPaperDir,"4b_Resample_GeneValues_keyPathsLPS_Down.png"), width = 4, height = 4)

#add color change to multi plot
lpsPlotMelt$emmremlStdBetaStatus <- c(lpsPlot$emmremlStdBetaStatus,lpsPlot$emmremlStdBetaStatus)

ggplot(data = lpsPlotMelt,
       aes(x = variable,
           y = value,
           group = geneName,
           col = emmremlStdBetaStatus > 0)) +
  geom_line(linewidth = 1) +
  theme_minimal() + theme_bw() + 
  theme(legend.position = "none") +
  ylab("Mean GE in Long. Samples") + xlab("Status") +
  scale_color_manual(values = c("#406F8B","#A3D65A")) +
  ggtitle("Key Pathway Genes Genes Sig DIFF in Dom F Meerkats",
          subtitle = paste0("Resampled GE Values, ",trt,", Treatment, FDR < 0.10"))
ggsave(paste0(figPaperDir,"4b_Resample_GeneValues_keyPathsLPS_All.png"), width = 6, height = 4)

```



### TODO ### - need to add macaques to the beta-beta comparison


```{r Figure 4 GSEA}
#tested pathways -- immune, signalling, pathway only
testedPaths <- unique(subset(gene.set, Category %in% c("signaling","pathway","immune","ctra") & GO.Term.Accession != "HALLMARK_INFLAMMASOME")$GO.Term.Accession)

pointSize <- 1
axisSize <- .25
labelSize <- 14
numberSize <- 10


allSsurGSEA <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/gsea_trtstatuspreg_GSEA_allResults.txt",
                      sep = ",", header = TRUE)

#didn't test Inflammasome in all spp
allSsurGSEA <- subset(allSsurGSEA, !V1 == "HALLMARK_INFLAMMASOME")

#bring in results from multi-spp analysis
allMultiSppGSEA <- read.table("~/Dropbox (Personal)/socialSignal/output/gsea_multiSPP_GSEA_allResults.txt", sep = ",", header = T)

colnames(allSsurGSEA)[c(1,4,5)] <- colnames(allMultiSppGSEA)[c(1,4,5)] <- c("pathway","ESup","ESdown")

ssurStatus <- subset(allSsurGSEA, trt %in% c("Ctrl","LPS") & c == "status")
#replace meerkat ES in all MultiSppGSEA

ssurStatus <- ssurStatus[,colnames(ssurStatus) != "c"]
#

#swap trt LPS plus and Ctrl minus
ssurStatus$trt <- ifelse(ssurStatus$trt == "LPS", "plus", "minus")
ssurStatus$shortName <- ifelse(ssurStatus$trt == "plus", "meerkatBetasFLPS_plusTrt", "meerkatBetasFLPS_minusTrt")

allMultiSppGSEA <- subset(allMultiSppGSEA, ! shortName %in% c("meerkatBetasFLPS_plusTrt","meerkatBetasFLPS_minusTrt"))

allMultiSppGSEA <- rbind.data.frame(allMultiSppGSEA,ssurStatus)

#limit to SGE1, Baboon, Meerkats
multi3SppGSEA <- subset(allMultiSppGSEA, shortName %in% c("BaboonMale_plusTrt","BaboonMale_minusTrt",
                                                          "BaboonFemale_plusTrt","BaboonFemale_minusTrt",
                                                          "SGE1_plusTrt","SGE1_minusTrt",
                                                          "meerkatBetasFLPS_plusTrt","meerkatBetasFLPS_minusTrt"))

multi3SppGSEA$species <- ifelse(multi3SppGSEA$shortName %in% c("BaboonMale_plusTrt","BaboonMale_minusTrt",
                                                          "BaboonFemale_plusTrt","BaboonFemale_minusTrt"), "Baboon",
                                ifelse(multi3SppGSEA$shortName %in% c("SGE1_plusTrt","SGE1_minusTrt"), "Macaque", "Meerkat"))

multi3SppGSEA$sex <- ifelse(multi3SppGSEA$species %in% c("Macaque","Meerkat"), "F",
                            ifelse(multi3SppGSEA$shortName %in% c("BaboonFemale_plusTrt","BaboonFemale_minusTrt"), "F", "M"))

multi3SppGSEA <- subset(multi3SppGSEA, pathway %in% testedPaths)

multi3SppGSEA$pvalUpAdj <- multi3SppGSEA$pval_upreg * 27
multi3SppGSEA$pvalDnAdj <- multi3SppGSEA$pval_downreg * 27

immunePathways <- c("HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                    "HALLMARK_INFLAMMATORY_RESPONSE",
                    "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                    "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                    "HALLMARK_COMPLEMENT",
                    "HALLMARK_CTRA_UP",
                    "HALLMARK_CTRA_DOWN")

multi3SppGSEAImmune <- subset(allMultiSppGSEA, pathway %in% immunePathways)

#for baseline and control
for (Trt in c("plus","minus")) {
  #compare every combo of species-sex, plot ES if EITHER is significant, all and immune
  for (sppA in c("Baboon_M","Baboon_F","Macaque_F","Meerkat_F")) {
    for (sppB in c("Baboon_M","Baboon_F","Macaque_F","Meerkat_F")) {
      if (sppA != sppB) {
        aSpp <- strsplit(sppA, "_")[[1]][1]
        aSex <- strsplit(sppA, "_")[[1]][2]
        bSpp <- strsplit(sppB, "_")[[1]][1]
        bSex <- strsplit(sppB, "_")[[1]][2]
        
        sigPaths <- unique(c(subset(multi3SppGSEA, species == aSpp & sex == aSex & trt == Trt & (pvalUpAdj < .05 | pvalDnAdj < .05))$pathway,
                             subset(multi3SppGSEA, species == bSpp & sex == bSex & trt == Trt & (pvalUpAdj < .05 | pvalDnAdj < .05))$pathway))
        
        aES <- vector()
        bES <- vector()
        aPval <- vector()
        bPval <- vector()
        
        for (p in sigPaths) {
          #get more sig data from A
          aUp <- ifelse(subset(multi3SppGSEA, species == aSpp & sex == aSex & trt == Trt & pathway == p)$pvalUpAdj <
                          subset(multi3SppGSEA, species == aSpp & sex == aSex & trt == Trt & pathway == p)$pvalDnAdj,
                        TRUE, FALSE)
          
          if (aUp) {
            aES <- c(aES,subset(multi3SppGSEA, species == aSpp & sex == aSex & trt == Trt & pathway == p)$ESup)
            aPval <- c(aPval,subset(multi3SppGSEA, species == aSpp & sex == aSex & trt == Trt & pathway == p)$pvalUpAdj)
          } else {
            aES <- c(aES,subset(multi3SppGSEA, species == aSpp & sex == aSex & trt == Trt & pathway == p)$ESdown)
            aPval <- c(aPval,subset(multi3SppGSEA, species == aSpp & sex == aSex & trt == Trt & pathway == p)$pvalDnAdj)
          }
          
          #get more sig data from B
          bUp <- ifelse(subset(multi3SppGSEA, species == bSpp & sex == bSex & trt == Trt & pathway == p)$pvalUpAdj <
                          subset(multi3SppGSEA, species == bSpp & sex == bSex & trt == Trt & pathway == p)$pvalDnAdj,
                        TRUE, FALSE)
          if (bUp) {
            bES <- c(bES,subset(multi3SppGSEA, species == bSpp & sex == bSex & trt == Trt & pathway == p)$ESup)
            bPval <- c(bPval,subset(multi3SppGSEA, species == bSpp & sex == bSex & trt == Trt & pathway == p)$pvalUpAdj)
          } else {
            bES <- c(bES,subset(multi3SppGSEA, species == bSpp & sex == bSex & trt == Trt & pathway == p)$ESdown)
            bPval <- c(bPval,subset(multi3SppGSEA, species == bSpp & sex == bSex & trt == Trt & pathway == p)$pvalDnAdj)
          }

          
        }
        
        gseaComp <- cbind.data.frame(sigPaths,aES,aPval,bES,bPval)
        gseaComp$sigSpp <- ifelse(gseaComp$aPval < .05 & gseaComp$bPval < .05, "both",
                                  ifelse(gseaComp$aPval < .05, paste0("sig",sppA), paste0("sig",sppB)))
        
        ggplot(gseaComp,
               aes(x = aES, 
                   y = bES,
                   col = sigSpp,
                   shape = sigPaths %in% immunePathways)) +
          geom_hline(yintercept = 0) +
          geom_vline(xintercept = 0) +
          geom_point(size = 3) +
          theme_bw() +
          labs(col = "p < 0.05",
               shape = "Immmune\nPathway") +
          xlab(paste0(sppA," ",Trt,"-LPS")) +
          ylab(paste0(sppB," ",Trt,"-LPS")) +
          ggtitle(paste0(Trt," Pathway ES - ",sppB," v. ",sppA))
        ggsave(paste0(figPaperDir,"GSEA/",Trt,"_",sppB,"v",sppA,".png"),
               height = 4, width = 6)
      }
    }
  }
}


### bar plot \ immune paths
multi3SppGSEAImmuneHist <- subset(multi3SppGSEA, pathway %in% immunePathways)

multi3SppGSEAImmuneHist$shortName <- factor(multi3SppGSEAImmuneHist$shortName, levels = c("meerkatBetasFLPS_plusTrt" ,"meerkatBetasFLPS_minusTrt",
                                                                                          "BaboonMale_plusTrt","BaboonMale_minusTrt",
                                                                                          "BaboonFemale_plusTrt","BaboonFemale_minusTrt",
                                                                                          "SGE1_plusTrt","SGE1_minusTrt"))

multi3SppGSEAImmuneHist$plotES <- ifelse(multi3SppGSEAImmuneHist$pvalUpAdj > .05 & multi3SppGSEAImmuneHist$pvalDnAdj > .05, 0,
                                         ifelse(multi3SppGSEAImmuneHist$pvalUpAdj < multi3SppGSEAImmuneHist$pvalDnAdj,
                                                multi3SppGSEAImmuneHist$ESup, multi3SppGSEAImmuneHist$ESdown))

ggplot(subset(multi3SppGSEAImmuneHist, trt == "plus"),
       aes(y = plotES,
           x = shortName,
           col = shortName)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linewidth = 2, col = "gray30") +
  facet_grid(cols = vars(str_replace_all(string = pathway, pattern = "_", replacement = "\n"))) +
  theme_bw() + theme(axis.text.x = element_blank())
  
ggsave(paste0(figPaperDir,"GSEA/plusLPS_ImmunePaths_dotplot.png"), height = 4, width = 6)

ggplot(subset(multi3SppGSEAImmuneHist, trt == "minus"),
       aes(y = plotES,
           x = shortName,
           col = shortName)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linewidth = 2, col = "gray30") +
  facet_grid(cols = vars(str_replace_all(string = pathway, pattern = "_", replacement = "\n"))) +
  theme_bw() + theme(axis.text.x = element_blank())
  
ggsave(paste0(figPaperDir,"GSEA/minusLPS_ImmunePaths_dotplot.png"), height = 4, width = 6)


### bar plot \ all pathways

multi3SppGSEA$plotES <- ifelse(multi3SppGSEA$pvalUpAdj > .05 & multi3SppGSEA$pvalDnAdj > .05, 0,
                                         ifelse(multi3SppGSEA$pvalUpAdj < multi3SppGSEA$pvalDnAdj,
                                                multi3SppGSEA$ESup, multi3SppGSEA$ESdown))

multi3SppGSEA$plotPval <- ifelse(multi3SppGSEA$pvalUpAdj < .05, multi3SppGSEA$pvalUpAdj,
                                 ifelse(multi3SppGSEA$pvalDnAdj < .05, multi3SppGSEA$pvalDnAdj, 1))

multi3SppGSEA$plotPval <- ifelse(multi3SppGSEA$plotPval == 0, .00001 * 27, multi3SppGSEA$plotPval)

multi3SppGSEA$shortName <- factor(multi3SppGSEA$shortName, levels = c("meerkatBetasFLPS_plusTrt" ,"meerkatBetasFLPS_minusTrt",
                                                                      "BaboonMale_plusTrt","BaboonMale_minusTrt",
                                                                      "BaboonFemale_plusTrt","BaboonFemale_minusTrt",
                                                                      "SGE1_plusTrt","SGE1_minusTrt"))

#which pathways have a sig hit
sigPathwaysMinus <- vector()
sigPathwaysPlus <- vector()

for (p in unique(multi3SppGSEA$pathway)) {
  minPathPplus <- min(c(subset(multi3SppGSEA, pathway == p & species == "Meerkat" & trt == "plus")$pvalUpAdj,
                        subset(multi3SppGSEA, pathway == p & species == "Meerkat" & trt == "plus")$pvalDnAdj))
  minPathPminus <- min(c(subset(multi3SppGSEA, pathway == p & species == "Meerkat" & trt == "minus")$pvalUpAdj,
                        subset(multi3SppGSEA, pathway == p & species == "Meerkat" & trt == "minus")$pvalDnAdj))
  if (minPathPplus < .05) {
    sigPathwaysPlus <- c(sigPathwaysPlus,p)
  }
  if (minPathPminus < .05) {
    sigPathwaysMinus <- c(sigPathwaysMinus,p)
  }
}

multi3SppGSEAplus <- subset(multi3SppGSEA, pathway %in% sigPathwaysPlus)
multi3SppGSEAminus <- subset(multi3SppGSEA, pathway %in% sigPathwaysMinus)


ggplot(subset(multi3SppGSEAplus, trt == "plus"),
       aes(y = plotES,
           x = shortName,
           col = shortName)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linewidth = 2, col = "gray30") +
  facet_grid(cols = vars(str_replace_all(string = pathway, pattern = "_", replacement = "\n"))) +
  theme_bw() + theme(axis.text.x = element_blank())
  
ggsave(paste0(figPaperDir,"GSEA/plusLPS_allSigPaths_dotplot_v2.png"), height = 4, width = 10)

ggplot(subset(multi3SppGSEAminus, trt == "minus"),
       aes(y = plotES,
           x = shortName,
           col = shortName)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linewidth = 2, col = "gray30") +
  facet_grid(cols = vars(str_replace_all(string = pathway, pattern = "_", replacement = "\n"))) +
  theme_bw() + theme(axis.text.x = element_blank())
  
ggsave(paste0(figPaperDir,"GSEA/minusLPS_allSigPaths_dotplot_v2.png"), height = 4, width = 6)


#compare betas - meerkat to baboon male
baboonMbetas <- read.csv2(paste0("~/Dropbox (Personal)/socialSignal/data/","male_baboon_all_emmreml_betas.csv"),sep = ",", dec = ".")
colnames(baboonMbetas)[1] <- "GeneID"

### FLIPPING ALL RANK BETAS ###
### RANK BETA IS NOW w.r.t. -rank ###
### SO RANK AGREES WITH MACAQUE Elo ###

baboonMbetas$MalerankbetaLPSminus <- -baboonMbetas$MalerankbetaLPSminus
baboonMbetas$MalerankbetaLPSplus <- -baboonMbetas$MalerankbetaLPSplus

rownames(baboonMbetas) <- baboonMbetas$GeneID

negVarGenes <- unique(c(rownames(baboonMbetas[which(baboonMbetas$MalerankvarbetaLPSplus < 0),]),
                        rownames(baboonMbetas[which(baboonMbetas$MalerankvarbetaLPSminus < 0),])))

baboonMbetas <- baboonMbetas[!rownames(baboonMbetas) %in% negVarGenes,]

baboonMbetas$stdBetaMalerankLPSminus <- baboonMbetas$MalerankbetaLPSminus / sqrt(baboonMbetas$MalerankvarbetaLPSminus)
baboonMbetas$stdBetaMalerankLPSplus <- baboonMbetas$MalerankbetaLPSplus / sqrt(baboonMbetas$MalerankvarbetaLPSplus)



## LPS
mrktBetas <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
mrktBetas$stdBeta_status <- mrktBetas$beta_status / sqrt(mrktBetas$var_beta_status)
mrktBetas$GeneID <- rownames(mrktBetas)
primBetas <- baboonMbetas

compBetas <- merge(mrktBetas,primBetas,by = "GeneID", suffixes = c(".mrkt",".prim"))

ggplot(data = compBetas, 
       aes(x = stdBeta_status,
           y = stdBetaMalerankLPSplus)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y~x")

#for each immune pathway
p <- "HALLMARK_INFLAMMATORY_RESPONSE"
for (p in immunePathways) {
  pathwayGenes <- unique(subset(gene.set, GO.Term.Accession == p)$Associated.Gene.Name)
  compBetaPath <- subset(compBetas, GeneID %in% pathwayGenes)
  
  plotR <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaMalerankLPSplus)$estimate, digits = 3)
  plotP <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaMalerankLPSplus)$p.value, digits = 3)
  
  ggplot(data = compBetaPath, 
         aes(y = stdBetaMalerankLPSplus,
             x = stdBeta_status)) +
    geom_point(size = pointSize) +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    geom_smooth(method = "lm", formula = "y~x") +
    theme_bw() +
    theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
    ggtitle(p, subtitle = paste0("R = ",plotR," | p = ",plotP)) +
    ylab(paste0("Stardized Effect Size:\nStatus in Male Baboons")) +
    xlab(paste0("Stardized Effect Size:\nStatus in Female Meerkats"))
  ggsave(paste0(figPaperDir,"GSEA/betas_LPS_MBab_v_Mrkt_",p,"_XYplot.png"), height = 4, width = 4)
  
  ## add "sign in either?"
  
  compBetaPath$SigInEither <- compBetaPath$fdr_status < .1 | compBetaPath$MalerankqvalueLPSplus < .1
  
  plotRsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                             compBetaPath$stdBetaMalerankLPSplus[compBetaPath$SigInEither])$estimate, digits = 3)
  plotPsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                             compBetaPath$stdBetaMalerankLPSplus[compBetaPath$SigInEither])$p.value, digits = 3)

  
  
  ggplot(data = compBetaPath, 
         aes(y = stdBetaMalerankLPSplus,
             x = stdBeta_status,
             col = SigInEither)) +
    geom_point() +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    geom_smooth(method = "lm", formula = "y~x") +
    theme_bw() +
    ggtitle(p,
            subtitle = paste0("Sig in Either: R = ",plotRsig," | p = ",plotPsig)) +
    ylab(paste0("Std Beta Rank Male Baboon LPS")) +
    xlab(paste0("Std Beta Status Female Meerkat LPS"))
    ggsave(paste0(figPaperDir,"GSEA/betas_LPS_MBab_v_Mrkt_",p,"_XYplot_wSig.png"), height = 4, width = 6)
}



## Ctrl
mrktBetas <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

mrktBetas$stdBeta_status <- mrktBetas$beta_status / sqrt(mrktBetas$var_beta_status)
mrktBetas$GeneID <- rownames(mrktBetas)
primBetas <- baboonMbetas

compBetas <- merge(mrktBetas,primBetas,by = "GeneID", suffixes = c(".mrkt",".prim"))

ggplot(data = compBetas, 
       aes(x = stdBeta_status,
           y = stdBetaMalerankLPSplus)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y~x")

#for each immune pathway
#p <- "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
for (p in immunePathways) {
  pathwayGenes <- unique(subset(gene.set, GO.Term.Accession == p)$Associated.Gene.Name)
  compBetaPath <- subset(compBetas, GeneID %in% pathwayGenes)
  
  plotR <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaMalerankLPSminus)$estimate, digits = 3)
  plotP <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaMalerankLPSminus)$p.value, digits = 3)
  
  ggplot(data = compBetaPath, 
         aes(y = stdBetaMalerankLPSminus,
             x = stdBeta_status)) +
    geom_point() +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    geom_smooth(method = "lm", formula = "y~x") +
    theme_bw() +
    ggtitle(p,
            subtitle = paste0("R = ",plotR," | p = ",plotP)) +
    ylab(paste0("Std Beta Rank Male Baboon Ctrl")) +
    xlab(paste0("Std Beta Status Female Meerkat Ctrl"))
  ggsave(paste0(figPaperDir,"GSEA/betas_Ctrl_MBab_v_Mrkt_",p,"_XYplot.png"), height = 4, width = 6)
  
  if (any(sum(compBetaPath$fdr_status < .1 | compBetaPath$MalerankqvalueLPSminus < .1))) {
    compBetaPath$SigInEither <- compBetaPath$fdr_status < .1 | compBetaPath$MalerankqvalueLPSminus < .1
    
    plotRsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaMalerankLPSminus[compBetaPath$SigInEither])$estimate, digits = 3)
    plotPsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaMalerankLPSminus[compBetaPath$SigInEither])$p.value, digits = 3)
  
    ggplot(data = compBetaPath, 
           aes(y = stdBetaMalerankLPSminus,
               x = stdBeta_status,
               col = SigInEither)) +
      geom_point() +
      geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
      geom_smooth(method = "lm", formula = "y~x") +
      theme_bw() +
      ggtitle(p,
              subtitle = paste0("Sig in Either: R = ",plotRsig," | p = ",plotPsig)) +
      ylab(paste0("Std Beta Rank Male Baboon Ctrl")) +
      xlab(paste0("Std Beta Status Female Meerkat Ctrl"))
      ggsave(paste0(figPaperDir,"GSEA/betas_Ctrl_MBab_v_Mrkt_",p,"_XYplot_wSig.png"), height = 4, width = 6)
  }
}



#compare betas - meerkat to baboon female
baboonFbetas <- read.csv2(paste0("~/Dropbox (Personal)/socialSignal/data/","female_baboon_all_emmreml_betas.csv"),sep = ",", dec = ".")
colnames(baboonFbetas)[1] <- "GeneID"

### FLIPPING ALL RANK BETAS ###
### RANK BETA IS NOW w.r.t. -rank ###
### SO RANK AGREES WITH MACAQUE Elo ###

baboonFbetas$FemalerankbetaLPSminus <- -baboonFbetas$FemalerankbetaLPSminus
baboonFbetas$FemalerankbetaLPSplus <- -baboonFbetas$FemalerankbetaLPSplus
baboonFbetas$FemaleDSIbetaLPSminus <- -baboonFbetas$FemaleDSIbetaLPSminus
baboonFbetas$FemaleDSIbetaLPSplus <- -baboonFbetas$FemaleDSIbetaLPSplus

rownames(baboonFbetas) <- baboonFbetas$GeneID

baboonFbetas$stdBetaFemalerankLPSminus <- baboonFbetas$FemalerankbetaLPSminus / sqrt(baboonFbetas$FemalerankvarbetaLPSminus)
baboonFbetas$stdBetaFemalerankLPSplus <- baboonFbetas$FemalerankbetaLPSplus / sqrt(baboonFbetas$FemalerankvarbetaLPSplus)

baboonFbetas$stdBetaFemaleDSILPSminus <- baboonFbetas$FemaleDSIbetaLPSminus / sqrt(baboonFbetas$FemaleDSIvarbetaLPSminus)
baboonFbetas$stdBetaFemaleDSILPSplus <- baboonFbetas$FemaleDSIbetaLPSplus / sqrt(baboonFbetas$FemaleDSIvarbetaLPSplus)

## LPS
mrktBetas <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
mrktBetas$stdBeta_status <- mrktBetas$beta_status / sqrt(mrktBetas$var_beta_status)
mrktBetas$GeneID <- rownames(mrktBetas)
primBetas <- baboonFbetas

compBetas <- merge(mrktBetas,primBetas,by = "GeneID", suffixes = c(".mrkt",".prim"))

ggplot(data = compBetas, 
       aes(x = stdBeta_status,
           y = stdBetaFemalerankLPSplus)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y~x")

#for each immune pathway
#p <- "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
for (p in immunePathways) {
  pathwayGenes <- unique(subset(gene.set, GO.Term.Accession == p)$Associated.Gene.Name)
  compBetaPath <- subset(compBetas, GeneID %in% pathwayGenes)
  
  plotR <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaFemalerankLPSplus)$estimate, digits = 3)
  plotP <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaFemalerankLPSplus)$p.value, digits = 3)
  
  ggplot(data = compBetaPath, 
         aes(y = stdBetaFemalerankLPSplus,
             x = stdBeta_status)) +
    geom_point() +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    geom_smooth(method = "lm", formula = "y~x") +
    theme_bw() +
    ggtitle(p,
            subtitle = paste0("R = ",plotR," | p = ",plotP)) +
    ylab(paste0("Std Beta Rank Female Baboon LPS")) +
    xlab(paste0("Std Beta Status Female Meerkat LPS"))
  ggsave(paste0(figPaperDir,"GSEA/betas_LPS_FBab_v_Mrkt_",p,"_XYplot.png"), height = 4, width = 6)
  
  if (sum(compBetaPath$fdr_status < .1 | compBetaPath$FemalerankqvalueLPSplus < .1) > 2) {
    compBetaPath$SigInEither <- compBetaPath$fdr_status < .1 | compBetaPath$FemalerankqvalueLPSplus < .1
    
    plotRsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaFemalerankLPSplus[compBetaPath$SigInEither])$estimate, digits = 3)
    plotPsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaFemalerankLPSplus[compBetaPath$SigInEither])$p.value, digits = 3)
  
    ggplot(data = compBetaPath, 
           aes(y = stdBetaFemalerankLPSplus,
               x = stdBeta_status,
               col = SigInEither)) +
      geom_point() +
      geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
      geom_smooth(method = "lm", formula = "y~x") +
      theme_bw() +
      ggtitle(p,
              subtitle = paste0("Sig in Either: R = ",plotRsig," | p = ",plotPsig)) +
      ylab(paste0("Std Beta Rank Female Baboon LPS")) +
      xlab(paste0("Std Beta Status Female Meerkat LPS"))
      ggsave(paste0(figPaperDir,"GSEA/betas_LPS_FBab_v_Mrkt_",p,"_XYplot_wSig.png"), height = 4, width = 6)
  }
}
  
## Ctrl
mrktBetas <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

mrktBetas$stdBeta_status <- mrktBetas$beta_status / sqrt(mrktBetas$var_beta_status)
mrktBetas$GeneID <- rownames(mrktBetas)
primBetas <- baboonFbetas

compBetas <- merge(mrktBetas,primBetas,by = "GeneID", suffixes = c(".mrkt",".prim"))

ggplot(data = compBetas, 
       aes(x = stdBeta_status,
           y = stdBetaFemalerankLPSminus)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y~x")

#for each immune pathway
p <- "HALLMARK_COMPLEMENT"
for (p in immunePathways) {
  pathwayGenes <- unique(subset(gene.set, GO.Term.Accession == p)$Associated.Gene.Name)
  compBetaPath <- subset(compBetas, GeneID %in% pathwayGenes)
  
  plotR <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaFemalerankLPSminus)$estimate, digits = 3)
  plotP <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaFemalerankLPSminus)$p.value, digits = 3)
  
  ggplot(data = compBetaPath, 
         aes(y = stdBetaFemalerankLPSminus,
             x = stdBeta_status)) +
    geom_point(size = pointSize) +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    geom_smooth(method = "lm", formula = "y~x") +
    theme_bw() +
    theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size = labelSize, color = "black"),
        axis.text = element_text(size = numberSize, color = "black")) +
    ggtitle(p, subtitle = paste0("R = ",plotR," | p = ",plotP)) +
    ylab(paste0("Stardized Effect Size:\nStatus in Female Baboons")) +
    xlab(paste0("Stardized Effect Size:\nStatus in Female Meerkats"))
  ggsave(paste0(figPaperDir,"GSEA/betas_Ctrl_FBab_v_Mrkt_",p,"_XYplot.png"), height = 4, width = 4)
  
  if (sum(compBetaPath$fdr_status < .1 | compBetaPath$FemalerankqvalueLPSminus < .1) > 2) {
    compBetaPath$SigInEither <- compBetaPath$fdr_status < .1 | compBetaPath$FemalerankqvalueLPSminus < .1
    
    plotRsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaFemalerankLPSminus[compBetaPath$SigInEither])$estimate, digits = 3)
    plotPsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaFemalerankLPSminus[compBetaPath$SigInEither])$p.value, digits = 3)
  
    ggplot(data = compBetaPath, 
           aes(y = stdBetaFemalerankLPSminus,
               x = stdBeta_status,
               col = SigInEither)) +
      geom_point() +
      geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
      geom_smooth(method = "lm", formula = "y~x") +
      theme_bw() +
      ggtitle(p,
              subtitle = paste0("Sig in Either: R = ",plotRsig," | p = ",plotPsig)) +
      ylab(paste0("Std Beta Rank Female Baboon Ctrl")) +
      xlab(paste0("Std Beta Status Female Meerkat Ctrl"))
      ggsave(paste0(figPaperDir,"GSEA/betas_Ctrl_FBab_v_Mrkt_",p,"_XYplot_wSig.png"), height = 4, width = 6)
  }
}


#compare betas - meerkat to macaque female
macaqueFbetas <- read.csv2(paste0("~/Dropbox (Personal)/socialSignal/data/","sge1_macaque_rankLPSall_emmreml_betas.csv"),sep = ",", dec = ".")
colnames(macaqueFbetas)[1] <- "GeneID"


rownames(macaqueFbetas) <- macaqueFbetas$GeneID

macaqueFbetas$stdBetaRankLPSminus <- macaqueFbetas$betaRankLPSminus / sqrt(macaqueFbetas$varBetaRankLPSminus)
macaqueFbetas$stdBetaRankLPSplus <- macaqueFbetas$betaRankLPSplus / sqrt(macaqueFbetas$varBetaRankLPSplus)


## LPS
mrktBetas <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
mrktBetas$stdBeta_status <- mrktBetas$beta_status / sqrt(mrktBetas$var_beta_status)
mrktBetas$GeneID <- rownames(mrktBetas)
primBetas <- macaqueFbetas

compBetas <- merge(mrktBetas,primBetas,by = "GeneID", suffixes = c(".mrkt",".prim"))

ggplot(data = compBetas, 
       aes(x = stdBeta_status,
           y = stdBetaRankLPSplus)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y~x")

#for each immune pathway
#p <- "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
for (p in immunePathways) {
  pathwayGenes <- unique(subset(gene.set, GO.Term.Accession == p)$Associated.Gene.Name)
  compBetaPath <- subset(compBetas, GeneID %in% pathwayGenes)
  
  plotR <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaRankLPSplus)$estimate, digits = 3)
  plotP <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaRankLPSplus)$p.value, digits = 3)
  
  ggplot(data = compBetaPath, 
         aes(x = stdBetaRankLPSplus,
             y = stdBeta_status)) +
    geom_point() +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    geom_smooth(method = "lm", formula = "y~x") +
    theme_bw() +
    ggtitle(p,
            subtitle = paste0("R = ",plotR," | p = ",plotP)) +
    xlab(paste0("Std Beta Rank Female Macaque LPS")) +
    ylab(paste0("Std Beta Status Female Meerkat LPS"))
  ggsave(paste0(figPaperDir,"GSEA/betas_LPS_Mrkt_v_FMacaque_",p,"_XYplot.png"), height = 4, width = 6)
  
  if (sum(compBetaPath$fdr_status < .1 | compBetaPath$fdrRankLPSplus < .1) > 2) {
    compBetaPath$SigInEither <- compBetaPath$fdr_status < .1 | compBetaPath$fdrRankLPSplus < .1
    
    plotRsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaRankLPSplus[compBetaPath$SigInEither])$estimate, digits = 3)
    plotPsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaRankLPSplus[compBetaPath$SigInEither])$p.value, digits = 3)
  
    ggplot(data = compBetaPath, 
           aes(x = stdBetaRankLPSplus,
               y = stdBeta_status,
               col = SigInEither)) +
      geom_point() +
      geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
      geom_smooth(method = "lm", formula = "y~x") +
      theme_bw() +
      ggtitle(p,
              subtitle = paste0("Sig in Either: R = ",plotRsig," | p = ",plotPsig)) +
      xlab(paste0("Std Beta Rank Female Macaque LPS")) +
      ylab(paste0("Std Beta Status Female Meerkat LPS"))
      ggsave(paste0(figPaperDir,"GSEA/betas_LPS_Mrkt_v_FMacaque_",p,"_XYplot_wSig.png"), height = 4, width = 6)
  }
}
  
## Ctrl
mrktBetas <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

mrktBetas$stdBeta_status <- mrktBetas$beta_status / sqrt(mrktBetas$var_beta_status)
mrktBetas$GeneID <- rownames(mrktBetas)
primBetas <- macaqueFbetas

compBetas <- merge(mrktBetas,primBetas,by = "GeneID", suffixes = c(".mrkt",".prim"))

ggplot(data = compBetas, 
       aes(x = stdBeta_status,
           y = stdBetaRankLPSminus)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y~x")

#for each immune pathway
p <- "HALLMARK_COMPLEMENT"
for (p in immunePathways) {
  pathwayGenes <- unique(subset(gene.set, GO.Term.Accession == p)$Associated.Gene.Name)
  compBetaPath <- subset(compBetas, GeneID %in% pathwayGenes)
  
  plotR <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaRankLPSminus)$estimate, digits = 3)
  plotP <- signif(cor.test(compBetaPath$stdBeta_status, compBetaPath$stdBetaRankLPSminus)$p.value, digits = 3)
  
  ggplot(data = compBetaPath, 
         aes(x = stdBetaRankLPSminus,
             y = stdBeta_status)) +
    geom_point() +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    geom_smooth(method = "lm", formula = "y~x") +
    theme_bw() +
    ggtitle(p,
            subtitle = paste0("R = ",plotR," | p = ",plotP)) +
    xlab(paste0("Std Beta Rank Female Macaque Ctrl")) +
    ylab(paste0("Std Beta Status Female Meerkat Ctrl"))
  ggsave(paste0(figPaperDir,"GSEA/betas_Ctrl_Mrkt_v_FMacaque_",p,"_XYplot.png"), height = 4, width = 6)
  
  if (sum(compBetaPath$fdr_status < .1 | compBetaPath$fdrRankLPSminus < .1) > 2) {
    compBetaPath$SigInEither <- compBetaPath$fdr_status < .1 | compBetaPath$fdrRankLPSminus < .1
    
    plotRsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaRankLPSminus[compBetaPath$SigInEither])$estimate, digits = 3)
    plotPsig <- signif(cor.test(compBetaPath$stdBeta_status[compBetaPath$SigInEither], 
                               compBetaPath$stdBetaRankLPSminus[compBetaPath$SigInEither])$p.value, digits = 3)
  
    ggplot(data = compBetaPath, 
           aes(x = stdBetaRankLPSminus,
               y = stdBeta_status,
               col = SigInEither)) +
      geom_point() +
      geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
      geom_smooth(method = "lm", formula = "y~x") +
      theme_bw() +
      ggtitle(p,
              subtitle = paste0("Sig in Either: R = ",plotRsig," | p = ",plotPsig)) +
      xlab(paste0("Std Beta Rank Female Macaque Ctrl")) +
      ylab(paste0("Std Beta Status Female Meerkat Ctrl"))
      ggsave(paste0(figPaperDir,"GSEA/betas_Ctrl_Mrkt_v_FMacaque_",p,"_XYplot_wSig.png"), height = 4, width = 6)
  }
}





xmasPlot <- subset(multi3SppGSEA, pvalUpAdj < .05 | pvalDnAdj < .05)
xmasPlot <- subset(xmasPlot, pathway %in% immunePathways)

xmasPlot$shortName <- factor(xmasPlot$shortName, levels = c("SGE1_minusTrt","SGE1_plusTrt",
                                                            "BaboonFemale_minusTrt","BaboonFemale_plusTrt",
                                                            "BaboonMale_minusTrt","BaboonMale_plusTrt",
                                                            "meerkatBetasFLPS_minusTrt","meerkatBetasFLPS_plusTrt"))

xmasPlot$pathway <- factor(xmasPlot$pathway, levels = c("HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                                        "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                                                        "HALLMARK_COMPLEMENT",
                                                        "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                                                        "HALLMARK_INFLAMMATORY_RESPONSE",
                                                        "HALLMARK_CTRA_UP",
                                                        "HALLMARK_CTRA_DOWN"))


#christmas tree plot
ggplot(data = xmasPlot, aes(x = shortName, y = pathway, col = plotES, size = -log10(plotPval))) +
  geom_point() + 
  #scale_color_viridis(limits = c(-1,1)) + 
  scale_color_gradientn(limits = c(-1,1),
                        colors = c(lowES,"grey80",highES)) +
  ggtitle("status - emreml - sig GSEA pathways") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(size = "-log10(adjP-Value)", col = "ES") +
   theme(legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(.5, 'cm'), #change legend key height
        legend.key.width = unit(.5, 'cm'), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))
ggsave(paste0(figPaperDir,"GSEA/status_plot_emreml_paper_draft.png"), width = 8, height = 6)

ggplot(data = subset(xmasPlot, pathway %in% c(c("HALLMARK_INTERFERON_GAMMA_RESPONSE",
                                                "HALLMARK_COMPLEMENT",
                                                "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                                                "HALLMARK_INFLAMMATORY_RESPONSE",
                                                "HALLMARK_CTRA_UP"))), aes(x = shortName, y = pathway, col = plotES, size = -log10(plotPval))) +
  geom_point() + 
  #scale_color_viridis(limits = c(-1,1)) + 
  scale_color_gradientn(limits = c(-1,1),
                        colors = c(lowES,"grey80",highES)) +
  ggtitle("status - emreml - sig GSEA pathways") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(size = "-log10(adjP-Value)", col = "ES") +
   theme(legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(.5, 'cm'), #change legend key height
        legend.key.width = unit(.5, 'cm'), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))
ggsave(paste0(figPaperDir,"GSEA/status_plot_emreml_paper_mrktOnly.png"), width = 8, height = 6)

ggplot(data = subset(xmasPlot, pathway %in% c(c("HALLMARK_INTERFERON_GAMMA_RESPONSE",
                                                "HALLMARK_COMPLEMENT",
                                                "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                                                "HALLMARK_INFLAMMATORY_RESPONSE"))), aes(x = shortName, y = pathway, col = plotES, size = -log10(plotPval))) +
  geom_point() + 
  #scale_color_viridis(limits = c(-1,1)) + 
  scale_color_gradientn(limits = c(-1,1),
                        colors = c(lowES,"grey80",highES)) +
  ggtitle("status - emreml - sig GSEA pathways") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(size = "-log10(adjP-Value)", col = "ES") +
   theme(legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(.5, 'cm'), #change legend key height
        legend.key.width = unit(.5, 'cm'), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))
ggsave(paste0(figPaperDir,"GSEA/status_plot_emreml_paper_mrktOnly_noCTRA.png"), width = 8, height = 6)

ggplot(data = subset(xmasPlot, pathway %in% c(c("HALLMARK_INTERFERON_GAMMA_RESPONSE",
                                                "HALLMARK_COMPLEMENT",
                                                "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                                                "HALLMARK_INFLAMMATORY_RESPONSE"))), aes(x = shortName, y = pathway, col = plotES, size = -log10(plotPval))) +
  geom_point() + 
  #scale_color_viridis(limits = c(-1,1)) + 
  scale_color_gradientn(limits = c(-1,1),
                        colors = c(lowES,"grey80",highES)) +
  ggtitle("") +
  theme_bw() +
  theme(title = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank()) +
  labs(size = "-log10(adjP-Value)", col = "ES") +
   theme(legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(.5, 'cm'), #change legend key height
        legend.key.width = unit(.5, 'cm'), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)); ggsave(paste0(figPaperDir,"GSEA/status_plot_emreml_poster.png"), width = 3.5, height = 2.5)

```


```{r Method Text}
pcaLPS <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_MF_LPS_10000genes_5lCPM_modelF_dfMeta.RDS")

LPSpc1r <- round(cor.test(as.numeric(pcaLPS$Trt != "Ctrl"), pcaLPS$PC1)$estimate, digits = 2)
LPSpc1p <- cor.test(as.numeric(pcaLPS$Trt != "Ctrl"), pcaLPS$PC1)$p.value
ggplot() +
  geom_point(aes(y = pcaLPS$PC1,
                 x = as.numeric(pcaLPS$Trt != "Ctrl")))


pcaGard <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_MF_Gard_10000genes_5lCPM_modelG_dfMeta.RDS")
Gardpc2r <- round(cor.test(as.numeric(pcaGard$Trt != "Ctrl"), pcaGard$PC2)$estimate, digits = 2)
Gardpc2p <- cor.test(as.numeric(pcaGard$Trt != "Ctrl"), pcaGard$PC2)$p.value

pcaDex <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_MF_Dex_10000genes_5lCPM_modelG_dfMeta.RDS")
Dexpc1r <- round(cor.test(as.numeric(pcaDex$Trt != "Ctrl"), pcaDex$PC1)$estimate, digits = 2)
Dexpc1p <- cor.test(as.numeric(pcaDex$Trt != "Ctrl"), pcaDex$PC1)$p.value
Dexpc2r <- round(cor.test(as.numeric(pcaDex$Trt != "Ctrl"), pcaDex$PC2)$estimate, digits = 2)
Dexpc2p <- cor.test(as.numeric(pcaDex$Trt != "Ctrl"), pcaDex$PC2)$p.value

print(paste0("We observed a strong gene regulatory response to stimulation in our samples. For all three stimulants, baseline samples separate from stimulated samples along the first or second principal component of variation in overall gene expression levels (rPC1-treatment condition=",LPSpc1r,", p=",LPSpc1p," for LPS, n=4,433 genes with logCPM > 5; rPC2-treatment condition=",Gardpc2r,", p=",Gardpc2p,"  for Gard, n=1,740 genes; rPC1-treatment condition=",Dexpc1r,", p=",Dexpc1p," and rPC2-treatment condition=",Dexpc2r,", p=",Dexpc2p," for Dex, n=4,663 genes;"))




fAllTrt <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_allTrt_10000genes_5lCPM_modelG_allTrtF+Preg+WeightwKmatrix_emmreml")
mAllTrt <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_allTrt_10000genes_5lCPM_modelG_allTrtM+WeightwKmatrix_emmreml")

fAllTrt$gene <- rownames(fAllTrt); mAllTrt$gene <- rownames(mAllTrt)

allAllTrt <- merge(fAllTrt, mAllTrt, by = "gene", suffixes = c(".f",".m"))

allStatR <- cor.test(allAllTrt$beta_status.f, allAllTrt$beta_status.m)$estimate
allStatP <- cor.test(allAllTrt$beta_status.f, allAllTrt$beta_status.m)$p.value

print(paste0("Further, estimates in males versus females are not well-correlated (Pearsons R=",allStatR,
             ", p=",allStatP,", Supplementary Figure x)."))

```

```{r Results Text}
allMetaF <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_allTrt_10000genes_5lCPM_metadata.RDS")
allMetaM <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_allTrt_10000genes_5lCPM_metadata.RDS")

allMeta <- rbind.data.frame(allMetaF,allMetaM)

print(paste0("RNA-sequencing in peripheral blood mononuclear cells (PBMCs) from ",
             length(unique(allMeta$correctsampleID))," individually recognized meerkats in the Kalahari study population (",
             length(unique(subset(allMeta, sex.long == "F")$correctsampleID))," females and ",
             length(unique(subset(allMeta, sex.long == "M")$correctsampleID))," males sampled from ",
             length(unique(allMeta$group))," social groups"))

captureMeta <- unique(cbind.data.frame(allMeta$captureref,allMeta$correctsampleID,allMeta$status_dom_sub.long,allMeta$sex,allMeta$avgWeight60d,allMeta$ageAtSample))
colnames(captureMeta) <- c("captureref","correctsampleID","status","sex","weight","age")

print(paste0("This data set combines samples from ",
             length(unique(subset(captureMeta, status == "D")$captureref))," animals who were dominant at the time of capture (",
             length(unique(subset(captureMeta, status == "D" & sex == "F")$captureref))," females and ",
             length(unique(subset(captureMeta, status == "D" & sex == "M")$captureref))," males) and ",
             length(unique(subset(captureMeta, status == "S")$captureref))," who were subordinate at the time of capture. ",
             sum(table(captureMeta$correctsampleID) > 1)," individuals were sampled repeatedly, ",
             length(unique(subset(allMeta, rationale == "newDom")$correctsampleID))," of whom are represented in our data set as both subordinates and dominants (i.e., were sampled across a dominance transition)."))

### age and weight avg and sd
captureMetaF <- subset(captureMeta, sex == "F")

paste0("Dominant females, including those in this study, tend to be heavier (dominant mean = ",
       mean(subset(captureMetaF,status == "D")$weight)," g, sd = ",sd(subset(captureMetaF,status == "D")$weight),
       "; subordinate mean = ",mean(subset(captureMetaF,status == "S")$weight)," g, sd = ",sd(subset(captureMetaF,status == "S")$weight),
       ") and older (dominant mean = ",mean(subset(captureMetaF,status == "D")$age)," yrs, sd = ",sd(subset(captureMetaF,status == "D")$age),
       "; subordinate mean = ",mean(subset(captureMetaF,status == "S")$age)," yrs, sd = ",sd(subset(captureMetaF,status == "S")$age),
       ", Clutton-Brock 2006, Kutsukake 2005) (Figure Sxxx).")

```

```{r }
totGenes <- 760
dirConsistent <- 715


print(paste0("For the ",totGenes,
             " gene-condition combinations we identified in the cross-sectional analysis (FDR < 10%; 735 unique genes, as 25 genes were dominance-associated in multiple conditions), ",
             dirConsistent," (",round(100 *dirConsistent / totGenes,digits = 0),
             "%) exhibited directionally consistent changes in mean expression levels across the dominance transition, (Figure X, p = ",bTest$p.value,")"))

```


```{r}
print(paste0("Sex differences are unlikely to result from differences in power, as our sample size for males (n=",
             length(unique(subset(allMeta, sex.long == "M")$correctsampleID))," unique males; n=",dim(subset(allMeta, sex.long == "M"))[1],
             " samples) exceeded our sample size for females (n=",
             length(unique(subset(allMeta, sex.long == "F")$correctsampleID))," unique females; n=",
             dim(subset(allMeta, sex.long == "F"))[1]," samples)."))
```

```{r}
pregSub <- length(unique(subset(allMetaF, pregStatusBackDate & status_dom_sub == "S")$captureref))

pregDomCapt <- unique(subset(allMetaF, pregStatusBackDate & status_dom_sub == "D")$captureref)
pregSubCapt <- unique(subset(allMetaF, pregStatusBackDate & status_dom_sub == "S")$captureref)


unique(paste0(subset(allMetaF, captureref %in% pregDomCapt)$captureref,"_",subset(allMetaF, captureref %in% pregDomCapt)$pregResult))
pregDomBorn <- 100*round(5/7, digits = 2)

unique(paste0(subset(allMetaF, captureref %in% pregSubCapt)$captureref,"_",subset(allMetaF, captureref %in% pregSubCapt)$pregResult))
pregSubBorn <- 100*round(1/7, digits = 2)


print(paste0("They are also more likely to be sampled when pregnant, even though subordinate females sometimes ",
             "conceive as well [cites]: in our sample, for example, ",pregSub,
             " subordinate samples were collected in pregnant females, although only ",pregSubBorn,
             "% of those pregnancies resulted in live offspring (compared to ",pregDomBorn,"% of pregnancies for dominant females)."))

length(unique(subset(allMetaF, pregStatusBackDate)$captureref))

length(unique(allMetaF$captureref))

length(unique(allMetaF$correctsampleID))

```

actually not true, need to find the slide where I implied this...
```{r}

immuneGenes <- unique(c(subset(gene.set, Category == c("immune","signaling"))$Associated.Gene.Name,
                        gene.set$Associated.Gene.Name[gene.set$GO.Term.Accession == "HALLMARK_TNFA_SIGNALING_VIA_NFKB"]))


mrktFLPSbetas <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
mrktFLPStrt <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_treatment")

lpsList <- rownames(mrktFLPSbetas)[mrktFLPSbetas$fdr_status < .1 & mrktFLPSbetas$beta_status > 0]
trtList <- rownames(mrktFLPStrt)[mrktFLPStrt$fdr_treatment < .1]

lpsTrtList <- intersect(lpsList,trtList)

mean(mrktFLPStrt$beta_treatment[rownames(mrktFLPStrt) %in% lpsList] > 0)

lpsImmuneList <- intersect(lpsList,immuneGenes)

mean(mrktFLPStrt$beta_treatment[rownames(mrktFLPStrt) %in% lpsImmuneList] > 0)

mrktFGardbetas <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Gardonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

mrktFGardtrt <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Gard_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml")


mrktLPSresp <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/",
                                 "allBatches_F_LPS_10000genes_5lCPM_modelF_basicResp+Preg+Weight_emmreml"))

mrktGardresp <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/",
                                 "allBatches_F_Gard_10000genes_5lCPM_modelG_basicResp+Preg+Weight_emmreml"))




print("In support of this possibility, all genes that are more highly expressed in dominant individuals in the LPS or Gard condition are also upregulated upon stimulation with LPS or Gard, respectively (xxx).")
```



```{r}

allMultiSppGSEA <- read.table("~/Dropbox (Personal)/socialSignal/output/gsea_multiSPP_GSEA_allResults.txt", sep = ",", header = T)



print(paste0("Like high-ranking male baboons, high status female meerkats tend to upregulate genes in NFkB-mediated and interferon-mediated defense pathways (baboons: ",
             "NFkB at LPS challenge p=",subset(allMultiSppGSEA, V1 == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" & shortName == "BaboonMale_plusTrt")$pval_upreg,
             " IFNg at baseline p=",subset(allMultiSppGSEA, V1 == "HALLMARK_INTERFERON_GAMMA_RESPONSE" & shortName == "BaboonMale_minusTrt")$pval_upreg,
             " meerkats: NFkB p=",subset(allMultiSppGSEA, V1 == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" & shortName == "meerkatBetasFLPS_plusTrt")$pval_upreg,
             " IFNg at baseline p=",signif(subset(allMultiSppGSEA, V1 == "HALLMARK_INTERFERON_GAMMA_RESPONSE" & shortName == "meerkatBetasFLPS_minusTrt")$pval_upreg, digits = 3),
             "), whereas the same pathways are downregulated in high-ranking female macaques (",
             "NFkB at LPS challenge p=",subset(allMultiSppGSEA, V1 == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" & shortName == "SGE1_plusTrt")$pval_downreg,
             " IFNg at baseline p=",signif(subset(allMultiSppGSEA, V1 == "HALLMARK_INTERFERON_GAMMA_RESPONSE" & shortName == "SGE1_minusTrt")$pval_downreg, digits = 3),
             ") and female baboons (",
             "NFkB at LPS challenge p=",subset(allMultiSppGSEA, V1 == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" & shortName == "BaboonFemale_plusTrt")$pval_downreg,
             " IFNg at baseline p=",subset(allMultiSppGSEA, V1 == "HALLMARK_INTERFERON_GAMMA_RESPONSE" & shortName == "BaboonFemale_minusTrt")$pval_downreg,
             ")."))

```

Reproductive Skew
```{r Reproductive Skew}
repSkew <- read.csv2(paste0(dataDir,"reprodSkew.csv"), sep = ",", dec = ".")

repSkew$SppSex <- paste0(repSkew$Species,"\n",repSkew$Sex)
repSkew$CommonName <- ifelse(repSkew$Species == "Papio_cynocephalus","baboon",
                             ifelse(repSkew$Species == "Suricatta_suricata","meerkat","macaque"))
repSkew$NameSex <- paste0(repSkew$CommonName,"\n",repSkew$Sex)


ggplot(repSkew,
       aes(x = NameSex,
           y = Standardized_variance_success)) +
  geom_point(size = 5, color = c(ctrlF,
                                 ctrlM,
                                 ctrlF,
                                 ctrlM,
                                 ctrlF,
                                 ctrlM)) +
  xlab("Species-Sex") +
  ylab("Standardized Variance of\nReproductive Success") +
  theme_minimal() + theme_bw()

ggsave(paste0(figPaperDir,"5_repSkew_LRSVariance.png"), width = 6, height = 4)

repSkewPlot <- subset(repSkew, NameSex %in% c("baboon\nmales",
                                              "baboon\nfemales",
                                              "meerkat\nfemales",
                                              "macaque\nfemales"))

repSkewPlot$NameSex <- factor(repSkewPlot$NameSex, levels = c("macaque\nfemales",
                                                              "baboon\nfemales",
                                                              "baboon\nmales",
                                                              "meerkat\nfemales"))

ggplot(repSkewPlot,
       aes(x = NameSex,
           y = Opposite_Sex_Ratio,
           col = log2(Opposite_Sex_Ratio))) +
  geom_point(size = 20) +
  geom_hline(yintercept = 1, col = "grey15") +
  xlab("Species-Sex") +
  ylab("Ratio LRS v. Opposite Sex") +
  theme_minimal() + theme_bw() + ylim(c(0,2)) +
  theme(axis.text.x = element_text(size = 14)) +
  scale_color_gradient(low = lowLRS, high = highLRS)

ggsave(paste0(figPaperDir,"5_repSkew_OppSexRatio.png"), width = 6, height = 4)
```

```{r percent of known birthdays}
read.csv2(paste0(dataDir,"KMPBirthdays.csv"), sep = ",")


```

```{r}

allSamples <- unique(metadata[,colnames(metadata) %in% c("captureref","correctsampleID","sex","rationale","ageAtSample","pregStatusBackDate35","group","status_dom_sub")])

allSamples <- subset(allSamples, ageAtSample > .95)

dim(allSamples)

length(unique(allSamples$correctsampleID))
length(unique(allSamples$group))



```


```{r gene list sig in response}
#nested emmreml run in genes only significant for status (711)
nestLPS <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPS_10000genes_5lCPM_modelF_nested+Preg+WeightwKmatrix_emmreml")

lpsOnly <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")
ctrlOnly <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")

#sig status
sigLPSCtrl <- c(rownames(ctrlOnly)[ctrlOnly$fdr_status < .1],
                rownames(lpsOnly)[lpsOnly$fdr_status < .1])

nestLPSsigOnly <- nestLPS[rownames(nestLPS) %in% sigLPSCtrl,]

#Status * Treatment Beta = Trt:Status Beta - Ctrl:Status Beta
#Status * Treatment Variance = sqrt((Trt:Status Var)^2 + (Ctrl:Status)^2)
#Status * Treatment P-Value = Use Std Beta as T.Test statistic, df=10,000

nestLPSsigOnly$beta_interaction <- nestLPSsigOnly$beta_Trt.Status - nestLPSsigOnly$beta_Ctrl.Status
nestLPSsigOnly$var_beta_interaction <- sqrt(nestLPSsigOnly$var_beta_Trt.Status^2 + nestLPSsigOnly$var_beta_Ctrl.Status^2)

nestLPSsigOnly$t_interaction <- nestLPSsigOnly$beta_interaction / sqrt(nestLPSsigOnly$var_beta_interaction)
nestLPSsigOnly$pval_interaction <- 2*pt(q=abs(nestLPSsigOnly$t_interaction), df=10000, lower.tail=FALSE)

ggplot() + geom_histogram(aes(x = nestLPSsigOnly$pval_interaction), bins = 100)


#sig trt
nestLPSsigOnly$qval_interaction <- qvalue(nestLPSsigOnly$pval_interaction)$qvalue

#26 genes
genesSigMag <- rownames(subset(nestLPSsigOnly, qval_interaction < .1))

nestLPSsigOnly[rownames(nestLPSsigOnly) %in% c("COMMD1","COMMD5","AKT1","EDN1"),]


```


### MYD88 and TRIF gene comparison, correlation bar chart
```{r fig 5 F}
paths <- rep(c("MYD88all","MYD88sig","TRIFall","TRIFsig"),3)
pathOnly <- rep(c("MYD88","MYD88","TRIF","TRIF"),3)
geneSet <- rep(c("all","sig","all","sig"),3)
spps <- c(rep("fMac",4),rep("fBab",4),rep("mBab",4))
pathwayGeneCorrs <- c(c(-.1,-.14,-.08,-.11), #fMac
                      c(-.25,-.31,-.16,-.34),  #fBab
                      c(.15,.29,.16,.17)) #mBab
sigNonSig <- c("ns","ns","ns","ns",
               "sig","ns","sig","ns",
               "ns","sig","sig","ns")
nGenes <- c(c(149,110,231,164), #fMac
            c(154,27,234,27),  #fBab
            c(154,80,234,97)) #mBab

allpathCorrs <- cbind.data.frame(spps,paths,geneSet,pathOnly,pathwayGeneCorrs,sigNonSig,nGenes)
allpathCorrs$sppPath <- paste0(allpathCorrs$spps,".",allpathCorrs$paths)

allpathCorrs$sppPath <- factor(allpathCorrs$sppPath, levels = allpathCorrs$sppPath[order(allpathCorrs$pathOnly)])

ggplot() +
  geom_point(data = allpathCorrs,
       aes(x = sppPath,
           y = pathwayGeneCorrs,
           col = sigNonSig,
           size = nGenes,
           shape = geneSet)) +
  geom_hline(yintercept = 0) +
  theme_minimal() + theme_bw() +
  scale_color_manual(values = c("gray40","firebrick3")) +
  scale_shape_manual(values = c(16,15)) +
  theme(axis.text.y = element_text(size = 16),
        axis.text.x = element_text(size = 4, angle = 90, vjust = 0.5),
        axis.title = element_blank())
ggsave(paste0(figPaperDir,"5g_TRIFMYD88_sppCorr.png"), width = 6, height = 4)


```

Calculate the statistical support for status with and without controlling for age, weight, pregnancy
```{r}
treatment <- "LPS"

for (treatment in c("Ctrl","LPS","Dex","Gard")) {
  #model Var is the variable to check the effect of before and after controlling for the control variable (ctrlVar)
  modelVar <- "Status"
  ctrlVar <- "Preg"
  
  #for (ctrlVar in c("Preg","Weight","StatusCentWeight")) {
  for (ctrlVar in c("Preg","Weight")) {
    
    modelTitle <- paste0("modeling ",ctrlVar," as Tech Var.")
    
    origModel <- "F"
    if (ctrlVar == "Preg") {
      betaTitle <- "preg"
      if (treatment %in% c("LPS","Ctrl")) {
        ctrlModel <- "P"
      } else {
        ctrlModel <- "Q"
        origModel <- "G"
      }
    } else if (ctrlVar == "Weight") {
      betaTitle <- "weight"
      if (treatment %in% c("LPS","Ctrl")) {
        ctrlModel <- "T"
      } else {
        origModel <- "G"
        ctrlModel <- "U"
      }
    } else if (ctrlVar == "StatusCentWeight") {
      betaTitle <- "weight"
      if (treatment %in% c("LPS","Ctrl")) {
        ctrlModel <- "V"
      } else {
        origModel <- "G"
        ctrlModel <- "W"
      }
    }
    
    #LPS/Ctrl
    #P: F + Preg
    #Dex/Guard
    #Q: G + Preg
    
    #R: F + Status
    #S: G + Status
    
    #T: F + Weight
    #U: G + Weight

    #V: F + Status Centered Weight
    #W: G + Status Centered Weight

    #Q control for preg directly
    #
    emremlOrigAll <- read.table(paste0(outDir,"allBatches_F_",
                                       treatment,"only_10000genes_5lCPM_model",
                                       origModel,"_basic+Preg+WeightwKmatrix_emmreml"))
    emremlCtrlAll <- read.table(paste0(outDir,"allBatches_F_",
                                       treatment,"only_10000genes_5lCPM_model",
                                       ctrlModel,"_basic+Preg+WeightwKmatrix_emmreml"))
    

    emremlOrigAll$geneID <- rownames(emremlOrigAll)
    emremlCtrlAll$geneID <- rownames(emremlCtrlAll)

    emremlAllComp <- merge(emremlOrigAll, emremlCtrlAll, by = "geneID", suffixes = c(".orig",".ctrl"))

    
    ### id the controlled column here, copy to "beta_ctrlVar"
    emremlAllComp$betaCtrlVar.orig <- emremlAllComp[,which(colnames(emremlAllComp) == paste0("beta_",betaTitle,".orig"))]
    emremlAllComp$betaCtrlVar.ctrl <- emremlAllComp[,which(colnames(emremlAllComp) == paste0("beta_",betaTitle,".ctrl"))]
    
    plotR <- signif(cor.test(emremlAllComp$betaCtrlVar.ctrl,emremlAllComp$betaCtrlVar.orig)$estimate, digits = 6)
    plotPval <- signif(cor.test(emremlAllComp$betaCtrlVar.ctrl,emremlAllComp$betaCtrlVar.orig)$p.value, digits = 6) 

    
    #want to compare the betas of modelVar
    ggplot(data = emremlAllComp,
           aes(x = betaCtrlVar.orig,
               y = betaCtrlVar.ctrl)) +
      geom_point(alpha = .3) +
      geom_abline(slope = 1, intercept = 0, col = "grey30") +
      xlab(paste0("Beta ",ctrlVar," Original Model")) +
      ylab(paste0("Beta ",ctrlVar," Control Model")) +
      theme_minimal() + theme_bw() +
      ggtitle(paste0(ctrlVar," Beta Comparison | ",modelTitle," - ",treatment," only"),
              subtitle = paste0("R = ",plotR," | p = ",plotPval))
    ggsave(paste0(suppFigDir,"ctrl_v_original_",ctrlVar,"Betas_",treatment,"_ctrl",ctrlVar,"_allGenes.png"),
           width = 6, height = 4)
    
    plotR <- signif(cor.test(emremlAllComp$beta_status.ctrl,emremlAllComp$beta_status.orig)$estimate, digits = 6)
    plotPval <- signif(cor.test(emremlAllComp$beta_status.ctrl,emremlAllComp$beta_status.orig)$p.value, digits = 6) 
    
    ggplot(data = emremlAllComp,
           aes(x = beta_status.orig,
               y = beta_status.ctrl)) +
      geom_point(alpha = .3) +
      geom_abline(slope = 1, intercept = 0, col = "grey30") +
      xlab("Beta Status Original Model") +
      ylab("Beta Status Control Model") +
      theme_minimal() + theme_bw() +
      ggtitle(paste0("Status Beta Comparison | ",modelTitle," - ",treatment," only"),
              subtitle = paste0("R = ",plotR," | p = ",plotPval))
    ggsave(paste0(suppFigDir,"ctrl_v_original_statusBetas_",treatment,"_ctrl",ctrlVar,"_allGenes.png"),
           width = 6, height = 4)
    
    ### id the controlled column here, copy to "pval_ctrlVar"
    emremlOrigAll$pvalCtrlVar <- emremlOrigAll[,which(colnames(emremlOrigAll) == paste0("pval_",betaTitle))]

    emremlCtrlAll$pvalCtrlVar <- emremlCtrlAll[,which(colnames(emremlCtrlAll) == paste0("pval_",betaTitle))]
    
    ### number and identity of genes significant at q < 0.10
    
    ##### get qvalues for the pvalues
    emremlAllComp$qval_status.ctrl <- qvalue(emremlAllComp$pval_status.ctrl)$qvalue
    emremlAllComp$qval_status.orig <- qvalue(emremlAllComp$pval_status.orig)$qvalue
    
    ##### get raw number differences (104 v 108), presumably lost
    origSig <- sum(emremlAllComp$qval_status.orig < .1)
    ctrlSig <- sum(emremlAllComp$qval_status.ctrl < .1)
    
    ##### get identity differences (N genes moved from sig - n.s.)
    totalDiff <- sum(emremlAllComp$qval_status.orig < .1 & emremlAllComp$qval_status.ctrl > .1) + sum(emremlAllComp$qval_status.orig > .1 & emremlAllComp$qval_status.ctrl < .1)
    
    
    #want to compare the QQ plots of modelVar
    results <- "emremlOrigAll"
    for (results in c("emremlOrigAll","emremlCtrlAll")) {
      emremlResults <- get(results)
      #sort by status
      emremlResults <- emremlResults[order(emremlResults$pval_status),]
      emremlResults$qqExpStatus <- -log10(sort(runif(length(rownames(emremlResults)), min = 0, max = 1)))
      emremlResults$qqStatus <- -log10(emremlResults$pval_status)
      
      #sort by preg
      emremlResults <- emremlResults[order(emremlResults$pvalCtrlVar),]
      emremlResults$qqExpCtrlVar <- -log10(sort(runif(length(rownames(emremlResults)), min = 0, max = 1)))
      emremlResults$qqCtrlVar <- -log10(emremlResults$pvalCtrlVar)
      
      assign(x = paste0(results), value = emremlResults)
    }
    
    ggplot() +
      geom_abline(slope = 1, intercept = 0) +
      geom_point(aes(x = emremlOrigAll$qqExpCtrlVar,
                     y = emremlOrigAll$qqCtrlVar), col = "grey") +
      geom_point(aes(x = emremlCtrlAll$qqExpCtrlVar,
                     y = emremlCtrlAll$qqCtrlVar), col = "dodgerblue") +
      xlab("Expected -log10(pval)") +
      ylab("Actual -log10(pval)") +
      theme_minimal() + theme_bw() +
      ggtitle(paste0("All Genes - ",ctrlVar," qq",
              subtitle = paste0(modelTitle," - ",treatment," only")))
    ggsave(paste0(suppFigDir,"ctrl_v_original_",treatment,"_plot",modelVar,"_ctrl",ctrlVar,"_allGenes.png"),
           width = 5, height = 5)
    
    ggplot() +
      geom_abline(slope = 1, intercept = 0) +
      geom_point(aes(x = emremlOrigAll$qqExpStatus,
                     y = emremlOrigAll$qqStatus), col = "grey") +
      geom_point(aes(x = emremlCtrlAll$qqExpStatus,
                     y = emremlCtrlAll$qqStatus), col = "dodgerblue") +
      xlab("Expected -log10(pval)") +
      ylab("Actual -log10(pval)") +
      theme_minimal() + theme_bw() +
      theme(plot.subtitle=element_text(size=8)) +
      ggtitle(paste0("All Genes - Status qq | ",dim(emremlOrigAll)[1]," genes"),
              subtitle = paste0(modelTitle," - ",treatment," only | nSig: orig=",origSig," ctrl=",ctrlSig," | total changes=",totalDiff))
    ggsave(paste0(suppFigDir,"ctrl_v_original_",treatment,"_plot",ctrlVar,"_ctrl",ctrlVar,"_allGenes.png"),
           width = 5, height = 5)
    
 
  }
}




```


### effect of cell composition, R between before and after status betas 
```{r}

```



### status centered mass
```{r}
treatment <- "LPS"

for (treatment in c("Ctrl","LPS","Dex","Gard")) {
  #model Var is the variable to check the effect of before and after controlling for the control variable (ctrlVar)
  modelVar <- "Status"
  ctrlVar <- "StatusCentWeight"
  

    modelTitle <- paste0("modeling ",ctrlVar," as Tech Var.")
    
    origModel <- "F"
    if (ctrlVar == "Preg") {
      betaTitle <- "preg"
      if (treatment %in% c("LPS","Ctrl")) {
        ctrlModel <- "P"
      } else {
        ctrlModel <- "Q"
        origModel <- "G"
      }
    } else if (ctrlVar == "Weight") {
      betaTitle <- "weight"
      if (treatment %in% c("LPS","Ctrl")) {
        ctrlModel <- "T"
      } else {
        origModel <- "G"
        ctrlModel <- "U"
      }
    } else if (ctrlVar == "StatusCentWeight") {
      betaTitle <- "weight"
      if (treatment %in% c("LPS","Ctrl")) {
        ctrlModel <- "V"
      } else {
        origModel <- "G"
        ctrlModel <- "W"
      }
    }


    #V: F + Status Centered Weight
    #W: G + Status Centered Weight

    #Q control for preg directly
    #
    emremlOrigAll <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/fullAnalysis/output/",
                                       "allBatches_F_",
                                       treatment,"only_10000genes_5lCPM_model",
                                       origModel,"_basicPreg+Weight_emmreml"))
    emremlCtrlAll <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/fullAnalysis/output/",
                                       "allBatches_F_",
                                       treatment,"only_10000genes_5lCPM_model",
                                       ctrlModel,"_basicPreg+Weight_emmreml"))

    emremlOrigAll$geneID <- rownames(emremlOrigAll)
    emremlCtrlAll$geneID <- rownames(emremlCtrlAll)

    emremlAllComp <- merge(emremlOrigAll, emremlCtrlAll, by = "geneID", suffixes = c(".orig",".ctrl"))

    
    ### id the controlled column here, copy to "beta_ctrlVar"
    emremlAllComp$betaCtrlVar.orig <- emremlAllComp[,which(colnames(emremlAllComp) == paste0("beta_",betaTitle,".orig"))]
    emremlAllComp$betaCtrlVar.ctrl <- emremlAllComp[,which(colnames(emremlAllComp) == paste0("beta_",betaTitle,".ctrl"))]
    
    plotR <- signif(cor.test(emremlAllComp$betaCtrlVar.ctrl,emremlAllComp$betaCtrlVar.orig)$estimate, digits = 6)
    plotPval <- signif(cor.test(emremlAllComp$betaCtrlVar.ctrl,emremlAllComp$betaCtrlVar.orig)$p.value, digits = 6) 

    
    #want to compare the betas of modelVar
    ggplot(data = emremlAllComp,
           aes(x = betaCtrlVar.orig,
               y = betaCtrlVar.ctrl)) +
      geom_point(alpha = .3) +
      geom_abline(slope = 1, intercept = 0, col = "grey30") +
      xlab(paste0("Beta ",ctrlVar," Original Model")) +
      ylab(paste0("Beta ",ctrlVar," Control Model")) +
      theme_minimal() + theme_bw() +
      ggtitle(paste0(ctrlVar," Beta Comparison | ",modelTitle," - ",treatment," only"),
              subtitle = paste0("R = ",plotR," | p = ",plotPval))
    ggsave(paste0(suppFigDir,"ctrl_v_original_",ctrlVar,"Betas_",treatment,"_ctrl",ctrlVar,"_allGenes.png"),
           width = 6, height = 4)
    
    plotR <- signif(cor.test(emremlAllComp$beta_status.ctrl,emremlAllComp$beta_status.orig)$estimate, digits = 6)
    plotPval <- signif(cor.test(emremlAllComp$beta_status.ctrl,emremlAllComp$beta_status.orig)$p.value, digits = 6) 
    
    ggplot(data = emremlAllComp,
           aes(x = beta_status.orig,
               y = beta_status.ctrl)) +
      geom_point(alpha = .3) +
      geom_abline(slope = 1, intercept = 0, col = "grey30") +
      xlab("Beta Status Original Model") +
      ylab("Beta Status Control Model") +
      theme_minimal() + theme_bw() +
      ggtitle(paste0("Status Beta Comparison | ",modelTitle," - ",treatment," only"),
              subtitle = paste0("R = ",plotR," | p = ",plotPval))
    ggsave(paste0(suppFigDir,"ctrl_v_original_statusBetas_",treatment,"_ctrl",ctrlVar,"_allGenes.png"),
           width = 6, height = 4)
    
    ### id the controlled column here, copy to "pval_ctrlVar"
    emremlOrigAll$pvalCtrlVar <- emremlOrigAll[,which(colnames(emremlOrigAll) == paste0("pval_",betaTitle))]

    emremlCtrlAll$pvalCtrlVar <- emremlCtrlAll[,which(colnames(emremlCtrlAll) == paste0("pval_",betaTitle))]


    #want to compare the QQ plots of modelVar
    results <- "emremlOrigAll"
    for (results in c("emremlOrigAll","emremlCtrlAll")) {
      emremlResults <- get(results)
      #sort by status
      emremlResults <- emremlResults[order(emremlResults$pval_status),]
      emremlResults$qqExpStatus <- -log10(sort(runif(length(rownames(emremlResults)), min = 0, max = 1)))
      emremlResults$qqStatus <- -log10(emremlResults$pval_status)
      
      #sort by preg
      emremlResults <- emremlResults[order(emremlResults$pvalCtrlVar),]
      emremlResults$qqExpCtrlVar <- -log10(sort(runif(length(rownames(emremlResults)), min = 0, max = 1)))
      emremlResults$qqCtrlVar <- -log10(emremlResults$pvalCtrlVar)
      
      assign(x = paste0(results), value = emremlResults)
    }
    
    ggplot() +
      geom_abline(slope = 1, intercept = 0) +
      geom_point(aes(x = emremlOrigAll$qqExpCtrlVar,
                     y = emremlOrigAll$qqCtrlVar), col = "grey") +
      geom_point(aes(x = emremlCtrlAll$qqExpCtrlVar,
                     y = emremlCtrlAll$qqCtrlVar), col = "dodgerblue") +
      xlab("Expected -log10(pval)") +
      ylab("Actual -log10(pval)") +
      theme_minimal() + theme_bw() +
      ggtitle(paste0("All Genes - ",ctrlVar," qq",
              subtitle = paste0(modelTitle," - ",treatment," only")))
    ggsave(paste0(suppFigDir,"ctrl_v_original_",treatment,"_plot",modelVar,"_ctrl",ctrlVar,"_allGenes.png"),
           width = 5, height = 5)
    
    ggplot() +
      geom_abline(slope = 1, intercept = 0) +
      geom_point(aes(x = emremlOrigAll$qqExpStatus,
                     y = emremlOrigAll$qqStatus), col = "grey") +
      geom_point(aes(x = emremlCtrlAll$qqExpStatus,
                     y = emremlCtrlAll$qqStatus), col = "dodgerblue") +
      xlab("Expected -log10(pval)") +
      ylab("Actual -log10(pval)") +
      theme_minimal() + theme_bw() +
      ggtitle("All Genes - Status qq",
              subtitle = paste0(modelTitle," - ",treatment," only"))
    ggsave(paste0(suppFigDir,"ctrl_v_original_",treatment,"_plot",ctrlVar,"_ctrl",ctrlVar,"_allGenes.png"),
           width = 5, height = 5)
    
 
  
}




```


Plot the effect of status in each treatment condition (a qq plot for Control, LPS, Dex, & Gard), to show that Dex supresses status effects

```{r qqplots for status within treated samples FEMALE}

#generate a dataframe, with columns: gene name, GEdivision, GEtile, qqActual, qqExpected
geneNames <- vector()
qqActual <- vector()
qqExpected <- vector()
trtCondition <- vector()

## LPS
for (t in c("LPS","Ctrl","Dex","Gard")) {
  if (t %in% c("LPS","Ctrl")) {
    model <- "F"
  } else {
    model <- "G"
  }
  
  emremlallbatch <- read.table(paste0("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_",t,"only_10000genes_5lCPM_model",model,"_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status"))
  
  emremlallbatchTrtSort <- emremlallbatch[order(emremlallbatch$pval_status),]
  
  #no quantile
  geneNames <- c(geneNames,rownames(emremlallbatchTrtSort))
  qqActual <- c(qqActual,-log10(emremlallbatchTrtSort$pval_status))
  #qqExpected <- c(qqExpected,-log10(1:length(emremlallbatch$pval_status) / length(emremlallbatch$pval_status)))
  qqExpected <- c(qqExpected,-log10(sort(runif(length(rownames(emremlallbatchTrtSort)), min = 0, max = 1))))
  trtCondition <- c(trtCondition,rep(t,length(rownames(emremlallbatchTrtSort))))
}


qqTreatment <- cbind.data.frame(geneNames,qqActual,qqExpected,trtCondition)

ggplot(data = qqTreatment, aes(x = qqExpected,
                               y = qqActual,
                               col = trtCondition)) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point() +
  scale_color_manual(values = c(ctrlF,dexF,gardF,lpsF)) +
  theme_minimal() + theme_bw() +
  ggtitle("Status Treatment-only p-value qqPlot",
          subtitle = paste0("Female, LPS, All Batches, min logCPM > 5, unique genes = ",length(unique(qqTreatment$geneNames)))) +
  ylab("-log10(pval) Observed") + xlab("-log10(pval) Expected (from flat dist.)")


ggsave(filename = paste0(suppFigDir,"Female_allTrt_allBatches__StatusTrtPval_byTrt.png"), width = 6, height = 6)

```

```{r}



"In support of this possibility, all genes that are more highly expressed in dominant individuals in the LPS or Gard condition are also upregulated upon stimulation with LPS or Gard, respectively (Fishers exact test p = xxx; Supplementary Table xxx)"
```




```{r preg and mass models}
#LPS/Ctrl
#P: F + Preg
#Dex/Guard
#Q: G + Preg

#R: F + Status
#S: G + Status

#T: F + Weight
#U: G + Weight

#V: F + Status Centered Weight
#W: G + Status Centered Weight

#Q control for preg directly
#
treatment <- "LPS"


emremlPreg <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_preg")



emremlOrig <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_status")


emremlMass <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_weight")
sum(emremlMass$fdr_weight < .1)
emremlMass <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_weight")
sum(emremlMass$fdr_weight < .1)
emremlMass <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Gardonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_weight")
sum(emremlMass$fdr_weight < .1)
emremlMass <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Dexonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_weight")
sum(emremlMass$fdr_weight < .1)


emremlPreg <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_preg")
sum(emremlPreg$fdr_preg < .1)
emremlMass <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Ctrlonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_preg")
sum(emremlPreg$fdr_preg < .1)
emremlMass <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Gardonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_preg")
sum(emremlPreg$fdr_preg < .1)
emremlMass <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_Dexonly_10000genes_5lCPM_modelG_basic+Preg+WeightwKmatrix_emmreml_wPermFDR_v1_preg")
sum(emremlPreg$fdr_preg < .1)



"Pregnancy status and body mass (but not age) are also associated with gene expression levels in our sample (n = xxx  xxx body mass associated genes; n=xxx  xxx pregnancy-associated genes, FDR < xxx, "

```


```{r}

"In addition to controlling for pregnancy and body mass in our main model, post hoc analyses showed that regressing out body mass or pregnancy from the gene expression data did not alter our findings for dominance rank (r for dominance effect betas in the original model versus after first removing body mass and pregnancy effects > 0.99, Supplemental Materials)."

emremlOrig <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelF_basic+Preg+WeightwKmatrix_emmreml")
emremlMass <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelT_basic+Preg+WeightwKmatrix_emmreml")
emremlPreg <- read.table("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_LPSonly_10000genes_5lCPM_modelP_basic+Preg+WeightwKmatrix_emmreml")

colnames(emremlMass) <- paste0(colnames(emremlMass),".mass")
colnames(emremlPreg) <- paste0(colnames(emremlPreg),".preg")

cor.test(emremlMass$beta_status.mass,emremlOrig$beta_status)
cor.test(emremlPreg$beta_status.preg,emremlOrig$beta_status)


```

```{r}

allMetaF <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_F_allTrt_10000genes_5lCPM_metadata.RDS")
allMetaM <- readRDS("~/Dropbox (Personal)/meerkats/RNA/meerkatGE/output/allBatches_M_allTrt_10000genes_5lCPM_metadata.RDS")

#capture only
allMetaF <- unique(allMetaF[colnames(allMetaF) %in% c("group","captureref","sex","status_dom_sub",
                                                      "ageAtSample","evidenceofTB.","draw_time","bloodVolmL",
                                                      "sampleDate","sampleYear","sampleMonth","rationale","transitionID",
                                                      "correctsampleID","birthdate","collectionRationale","climSeason")])
allMetaF$sMonthDec <- round(as.numeric(format(as.Date(allMetaF$sampleDate, format="%Y-%m-%d"),"%m")) +
                              as.numeric(format(as.Date(allMetaF$sampleDate, format="%Y-%m-%d"),"%d")) / 31, digits = 2)

allMetaM <- unique(allMetaM[colnames(allMetaM) %in% c("group","captureref","sex","status_dom_sub",
                                                      "ageAtSample","evidenceofTB.","draw_time","bloodVolmL",
                                                      "sampleDate","sampleYear","sampleMonth","rationale","transitionID",
                                                      "correctsampleID","birthdate","collectionRationale","climSeason")])
allMetaM$sMonthDec <- round(as.numeric(format(as.Date(allMetaM$sampleDate, format="%Y-%m-%d"),"%m")) +
                              as.numeric(format(as.Date(allMetaM$sampleDate, format="%Y-%m-%d"),"%d")) / 31, digits = 2)


allMeta <- rbind.data.frame(allMetaF,allMetaM)




"Following quality control, our final data set included RNA-seq profiles from 113 individually recognized meerkats in the Kalahari study population (740 total samples from 52 females and 61 males in 15 social groups, obtained across xxx blood draws; Figure 1; Supplementary Table S1; mean = 5.87 million  5.06 million s.d reads; Supplementary Table S2)."


```

